%============================================================
%  app_quantum.tex
%  Appendix E: Operators on Hilbert Spaces — Mathematical
%  Foundations for Quantum Mechanics
%
%  Tensor Formats in Banach Spaces
%  Antonio Falcó, CEU Universities
%
%  This appendix develops the operator-theoretic vocabulary
%  needed for Part V (Chapters 12 and 13). The setting is
%  throughout a separable complex Hilbert space of arbitrary
%  (possibly infinite) dimension. Standard mathematical
%  notation is used exclusively; the Dirac bra-ket notation
%  is not used (see Remark~\ref{rem:app:dirac}).
%
%  Estimated length: ~22 pp.
%============================================================

\chapter{Operators on Hilbert Spaces: Mathematical Background}
\label{app:quantum}

This appendix collects the operator-theoretic foundations needed for
Part~V (Chapters~12 and~13). The presentation is self-contained,
mathematically rigorous, and set throughout in a separable complex
Hilbert space $H$ of \emph{arbitrary dimension} (finite or infinite).
No prior knowledge of quantum mechanics is assumed. Standard references
are~\cite{ReedSimon1972} and~\cite{BratteliRobinson1987}; for
Schatten ideals and their tensor-product interpretation
see~\cite{Ryan2002}.

%------------------------------------------------------------
\section{Complex Hilbert spaces}
\label{app:quantum:hilbert}
%------------------------------------------------------------

\subsection*{Inner products and Hilbert spaces}

A \emph{complex Hilbert space} is a complex vector space $H$ equipped
with an \emph{inner product} $\langle \cdot, \cdot \rangle: H \times H
\to \CC$, sesquilinear (conjugate-linear in the first argument, linear
in the second), Hermitian ($\langle u,v\rangle = \overline{\langle v,u\rangle}$),
and positive definite ($\langle u,u\rangle > 0$ for $u \neq 0$).
The induced norm $\|u\| \coloneqq \sqrt{\langle u,u\rangle}$ makes $H$
a normed space; when complete it is a \emph{Hilbert space}.

\begin{remark}[Convention]
\label{rem:app:sesqui}
We adopt the convention standard in mathematics: the inner product is
linear in the \emph{second} argument. Physics texts often reverse this.
\end{remark}

Throughout this appendix, $H$ denotes a \emph{separable} complex
Hilbert space, i.e., one admitting a countable orthonormal basis.
No finite-dimensionality is assumed unless explicitly stated.

\begin{example}
\label{ex:app:l2}
The space $\ell^2(\NN; \CC) = \{(a_n)_{n \geq 1} \subset \CC :
\sum_n |a_n|^2 < \infty\}$ with
$\langle a, b\rangle = \sum_n \overline{a_n} b_n$
is an infinite-dimensional separable Hilbert space. More generally,
$L^2(\Omega, \mu; \CC)$ for any $\sigma$-finite measure space $(\Omega,\mu)$
with $L^2(\Omega,\mu;\CC)$ separable is a Hilbert space. These are the
component spaces that arise in multipartite quantum systems.
\end{example}

\subsection*{Orthonormal bases and Parseval's identity}

A countable family $\{e_k\}_{k \geq 1}$ is an \emph{orthonormal basis}
of $H$ if $\langle e_j, e_k\rangle = \delta_{jk}$ and every $u \in H$
has the norm-convergent expansion
\begin{equation}
\label{eq:app:fourier}
    u = \sum_{k=1}^\infty \langle e_k, u\rangle\, e_k,
    \qquad
    \|u\|^2 = \sum_{k=1}^\infty |\langle e_k, u\rangle|^2
    \quad \text{(Parseval)}.
\end{equation}
All separable infinite-dimensional Hilbert spaces over $\CC$ are
isometrically isomorphic to $\ell^2(\NN;\CC)$.

\subsection*{The Riesz isomorphism and the dual $H^*$}

The topological dual of $H$ is $H^* \coloneqq \cL(H, \CC)$, equipped
with the operator norm. By the Riesz representation theorem, the map
\begin{equation}
\label{eq:app:riesz}
    J : H \longrightarrow H^*, \qquad
    J(v) \coloneqq \langle v, \cdot\rangle,
\end{equation}
is a conjugate-linear isometric isomorphism (i.e., $J(\lambda v)
= \overline{\lambda} J(v)$ and $\|J(v)\|_{H^*} = \|v\|_H$).
Thus $H$ and $H^*$ are isometrically isomorphic as real Banach spaces,
but not as complex Banach spaces (the isomorphism is conjugate-linear,
not linear).

\begin{remark}[Dual vs.\ conjugate space]
\label{rem:app:dual_vs_conjugate}
The \emph{conjugate space} $\overline{H}$ is defined as $H$ with the
same additive structure but scalar multiplication twisted by conjugation:
$\lambda \cdot_{\overline{H}} v \coloneqq \overline{\lambda} \cdot_H v$.
The Riesz map $J: H \to H^*$ is a conjugate-linear isomorphism, and so
is the canonical map $H \to \overline{H}$, $v \mapsto \bar v$.
Hence $H^*$ and $\overline{H}$ are \emph{canonically isomorphic as
complex Banach spaces} via $J(v) \leftrightarrow \bar v$.
However, the identification $H^* \cong \overline{H}$ is
\emph{not canonical in the absence of a chosen inner product}: $H^*$ is
defined purely in terms of the norm, whereas $\overline{H}$ uses the
linear structure. In this appendix we work with $H^*$ (the topological
dual) as the primary object, which is the correct setting for
infinite-dimensional tensor products; see
Remarks~\ref{rem:app:hs_tensor} and~\ref{rem:app:trace_projective}.
\end{remark}

\begin{remark}[Duality mapping vs.\ Riesz isomorphism]
\label{rem:app:riesz_duality}
In the real Banach space setting of Chapter~\ref{chap:banach}, the
duality mapping $J: X \to 2^{X^*}$ is set-valued and nonlinear in
general. In a complex Hilbert space, $J: H \to H^*$ is single-valued
and conjugate-linear. This simplification is the reason why
best-approximation and projection problems are easier in Hilbert spaces:
every closed subspace is complemented via orthogonal projection, and the
metric projection onto a closed convex set is linear when the set is a
subspace.
\end{remark}

\begin{remark}[Dirac notation]
\label{rem:app:dirac}
In the physics literature, elements of $H$ are written $|v\rangle$
(\emph{kets}) and elements of $H^*$ are written $\langle v|$
(\emph{bras}), with $\langle v| = J(v) \in H^*$ under our convention.
The inner product $\langle u, v\rangle_H$ becomes $\langle u|v\rangle$.
This notation, introduced by Dirac~\cite{Dirac1930}, is compact for
computation. This monograph uses exclusively standard mathematical
notation: $\langle u, v\rangle$ for the inner product, $J(v) \in H^*$
for the Riesz dual of $v$, and $u \otimes J(v) \in H \otimes H^*$
for rank-one operators.
\end{remark}

%------------------------------------------------------------
\section{Bounded operators and the adjoint}
\label{app:quantum:bounded}
%------------------------------------------------------------

\subsection*{The C*-algebra $\cL(H)$}

The space $\cL(H)$ of bounded $\CC$-linear operators $A: H \to H$,
with operator norm $\|A\| \coloneqq \sup_{\|u\|\leq 1}\|Au\|$, is a
Banach algebra satisfying $\|AB\| \leq \|A\|\|B\|$. It is moreover a
\emph{C*-algebra}: every $A \in \cL(H)$ has a unique \emph{adjoint}
$A^* \in \cL(H)$ characterised by
\begin{equation}
\label{eq:app:adjoint}
    \langle A^* u, v\rangle = \langle u, Av\rangle
    \quad \forall\, u, v \in H,
\end{equation}
satisfying $\|A^*\| = \|A\|$ and $\|A^*A\| = \|A\|^2$ (C*-identity).
The adjoint is the operator-theoretic analogue of the dual map
$A^T: Y^* \to X^*$ from Banach space theory, adapted to the inner
product structure: $A^* = J^{-1} \circ A^T \circ J$ where $A^T$ is
the Banach-space transpose.

\subsection*{Special classes of operators}

\begin{definition}
\label{def:app:operator_classes}
An operator $A \in \cL(H)$ is:
\begin{enumerate}
\item[\textup{(i)}] \emph{Self-adjoint} if $A = A^*$.
\item[\textup{(ii)}] \emph{Positive} ($A \geq 0$ or $A \succeq 0$)
    if $\langle u, Au\rangle \geq 0$ for all $u \in H$.
\item[\textup{(iii)}] \emph{Strictly positive} ($A > 0$ or $A \succ 0$)
    if $\langle u, Au\rangle > 0$ for all $u \neq 0$.
\item[\textup{(iv)}] \emph{Unitary} if $A^*A = AA^* = I$.
\item[\textup{(v)}] An \emph{orthogonal projection} if $A^2 = A = A^*$.
\end{enumerate}
\end{definition}

Every positive operator $A \geq 0$ has a unique positive square root
$A^{1/2} \geq 0$, and its \emph{absolute value} is
$|A| \coloneqq (A^*A)^{1/2}$.

\begin{proposition}[Orthogonal projections and closed subspaces]
\label{prop:app:proj_subspace}
There is a bijection between closed subspaces $S \subset H$ and
orthogonal projections $\PS{S} \in \cL(H)$, given by $S = \mathrm{Im}(\PS{S})$.
For every closed subspace $S$, the orthogonal decomposition
$H = S \oplus S^\perp$ holds, where $S^\perp = \ker(\PS{S})$.
\end{proposition}

In particular, \emph{every closed subspace of a Hilbert space is
complemented} (Proposition~\ref{prop:finitedim_split} gives this for
finite-dimensional subspaces of general Banach spaces; in Hilbert spaces
it holds for all closed subspaces). This is the key distinction between
Hilbert spaces and general Banach spaces noted in
Chapter~\ref{chap:banach}.

\subsection*{Rank-one operators}

For $u \in H$ and $\varphi \in H^*$, the \emph{rank-one operator}
$u \otimes \varphi \in \cL(H)$ is defined by
\begin{equation}
\label{eq:app:rank_one}
    (u \otimes \varphi)(w) \coloneqq \varphi(w)\, u, \quad w \in H,
\end{equation}
with norm $\|u \otimes \varphi\| = \|u\|_H \|\varphi\|_{H^*}$.
When $\varphi = J(v)$ for some $v \in H$, equation~\eqref{eq:app:rank_one}
gives $(u \otimes J(v))(w) = \langle v, w\rangle\, u$.
The orthogonal projection onto $\spann\{u\}$ (for $\|u\|=1$) is
\begin{equation}
\label{eq:app:rank_one_proj}
    \PS{\spann\{u\}} = u \otimes J(u).
\end{equation}

\begin{remark}[Rank-one operators and the tensor product $H \otimes H^*$]
\label{rem:app:rank_one_tensor}
The space of finite-rank operators is the algebraic tensor product
$H \otimes_a H^*$ embedded in $\cL(H)$ via~\eqref{eq:app:rank_one}.
This embedding is the starting point for the identification of Schatten
ideals with completions of $H \otimes_a H^*$ under appropriate tensor
norms; see Sections~\ref{app:quantum:hs} and~\ref{app:quantum:trace}.
Note that $H^*$ (not $H$) is the correct second factor: the map
$(u, \varphi) \mapsto u \otimes \varphi$ is \emph{bilinear} (linear in
both arguments), whereas $(u, v) \mapsto u \otimes J(v)$ is
sesquilinear. Working with $H^*$ therefore preserves the algebraic
tensor product structure of Chapter~\ref{chap:algtensor}.
\end{remark}

%------------------------------------------------------------
\section{Compact operators and singular values}
\label{app:quantum:compact}
%------------------------------------------------------------

\subsection*{Compact operators}

An operator $A \in \cL(H)$ is \emph{compact} if the image of the
unit ball under $A$ is relatively compact. The set $\cK(H) \subset \cL(H)$
of compact operators is a closed two-sided ideal. In a separable Hilbert
space, $\cK(H)$ is the norm-closure of the finite-rank operators.

\begin{theorem}[Spectral theorem for compact self-adjoint operators]
\label{thm:app:spectral}
Let $A \in \cK(H)$ be self-adjoint. Then there exist an orthonormal
sequence $\{u_k\}$ in $H$ and real scalars $\lambda_k \to 0$ such that
\begin{equation}
\label{eq:app:spectral}
    A = \sum_{k=1}^\infty \lambda_k\, u_k \otimes J(u_k),
\end{equation}
with convergence in operator norm. The $\lambda_k$ are the non-zero
eigenvalues of $A$ with eigenvectors $u_k$.
\end{theorem}

\subsection*{Singular values}

For $A \in \cK(H)$, the operator $|A| = (A^*A)^{1/2}$ is compact,
positive, and self-adjoint. Its eigenvalues, denoted $s_1(A) \geq s_2(A)
\geq \cdots \geq 0$ and called the \emph{singular values} of $A$,
satisfy $s_k(A) \to 0$.

\begin{theorem}[Singular value decomposition]
\label{thm:app:svd}
For $A \in \cK(H)$ with non-zero singular values $(s_k(A))_{k \geq 1}$
there exist orthonormal sequences $\{u_k\}$ and $\{v_k\}$ in $H$ and
functionals $\varphi_k = J(u_k) \in H^*$ such that
\begin{equation}
\label{eq:app:svd}
    A = \sum_{k=1}^\infty s_k(A)\, v_k \otimes J(u_k),
\end{equation}
with norm convergence. Here $\{u_k\}$ are eigenvectors of $A^*A$ and
$\{v_k\}$ of $AA^*$, so $Au_k = s_k(A) v_k$.
\end{theorem}

\begin{remark}[SVD and Tucker rank]
\label{rem:app:svd_tucker}
For $d=2$ and Hilbert spaces $H_1$, $H_2$, the SVD~\eqref{eq:app:svd}
is the Schmidt decomposition of a tensor in $H_1 \otimes_H H_2$
(Theorem~\ref{thm:app:hilbert_tensor_norms} and
Proposition~\ref{prop:app:hs_rank}). The number of non-zero singular
values is the Tucker rank $r_1(v) = r_2(v) = r$.
\end{remark}

%------------------------------------------------------------
\section{Hilbert--Schmidt operators}
\label{app:quantum:hs}
%------------------------------------------------------------

\subsection*{Definition and basic properties}

\begin{definition}
\label{def:app:hs}
An operator $A \in \cL(H)$ is \emph{Hilbert--Schmidt} if
\begin{equation}
\label{eq:app:hs_norm}
    \|A\|_2^2 \coloneqq \sum_{k=1}^\infty \|A e_k\|^2 < \infty
\end{equation}
for some (equivalently every) orthonormal basis $\{e_k\}$. The space
of Hilbert--Schmidt operators is $\cT{2}(H)$, with $\|A\|_2 =
(\Tr(A^*A))^{1/2} = (\sum_k s_k(A)^2)^{1/2}$.
\end{definition}

\begin{proposition}
\label{prop:app:hs_properties}
$\cT{2}(H)$ satisfies:
\begin{enumerate}
\item[\textup{(i)}] It is a Hilbert space under
    $\langle A, B\rangle_{\mathrm{HS}} \coloneqq \Tr(A^*B)
    = \sum_k \langle Ae_k, Be_k\rangle$.
\item[\textup{(ii)}] $\cT{2}(H) \subset \cK(H)$ and
    $\|A\| \leq \|A\|_2$.
\item[\textup{(iii)}] $\cT{2}(H)$ is a two-sided ideal:
    $\|BA\|_2 \leq \|B\|\|A\|_2$ for $B \in \cL(H)$.
\item[\textup{(iv)}] $A \in \cT{2}(H)$ iff $A^* \in \cT{2}(H)$,
    with $\|A^*\|_2 = \|A\|_2$.
\end{enumerate}
\end{proposition}

\subsection*{Tensor product identification}

\begin{theorem}
\label{thm:app:hs_tensor}
There is a canonical isometric isomorphism of Hilbert spaces
\begin{equation}
\label{eq:app:hs_tensor_iso}
    \cT{2}(H) \cong H \otimes_H H^*,
\end{equation}
where $H \otimes_H H^*$ denotes the Hilbert tensor product of $H$ and
$H^*$ \textup{(}with the inner product on $H^*$ induced by $J$:
$\langle J(u), J(v)\rangle_{H^*} \coloneqq \langle u, v\rangle_H$\textup{)}.
Explicitly, the elementary tensor $u \otimes \varphi \in H \otimes_a H^*$
corresponds under~\eqref{eq:app:hs_tensor_iso} to the rank-one operator
$w \mapsto \varphi(w)\, u$.
\end{theorem}

\begin{remark}[Why $H^*$ and not $\overline{H}$]
\label{rem:app:hs_tensor}
In finite dimensions $H^* \cong H$ (both isomorphic to $\CC^n$), so
the distinction between $H \otimes H^*$ and $H \otimes H$ is invisible.
In infinite dimensions $H$ and $H^*$ are not linearly isomorphic as
complex Banach spaces (the Riesz map is conjugate-linear, not linear).
Writing $\cT{2}(H) \cong H \otimes_H H^*$ is correct: the bilinear map
$(u, \varphi) \mapsto (u \otimes \varphi)$ is $\CC$-bilinear and extends
to the tensor product. Writing $\cT{2}(H) \cong H \otimes_H \overline{H}$
is also seen in the literature (since $\overline{H} \cong H^*$ as Banach
spaces via $J$), but introduces a conjugate-linear identification that
obscures the purely algebraic tensor structure. In this monograph we use
$H \otimes H^*$ to remain consistent with the Banach-space tensor
framework of Chapter~\ref{chap:algtensor}.
\end{remark}

\begin{remark}[Hilbert--Schmidt operators as Tucker rank-$(r,r)$ tensors]
\label{rem:app:hs_tucker}
Under~\eqref{eq:app:hs_tensor_iso}, a Hilbert--Schmidt operator of rank
$r$ corresponds to a tensor in $H \otimes_H H^*$ of Tucker rank $(r,r)$
(Proposition~\ref{prop:app:hs_rank} in Appendix~\ref{app:functional}).
Its SVD~\eqref{eq:app:svd} is the Tucker decomposition:
\begin{equation*}
    A = \sum_{k=1}^r s_k(A)\, v_k \otimes J(u_k)
    \in H \otimes_H H^*,
\end{equation*}
with $\Umin{1}{A} = \overline{\mathrm{Im}(A^*)} \subset H$ and
$\Umin{2}{A} = J(\overline{\mathrm{Im}(A)}) \subset H^*$ the minimal
subspaces in the two directions.
\end{remark}

%------------------------------------------------------------
\section{Trace-class operators}
\label{app:quantum:trace}
%------------------------------------------------------------

\subsection*{Definition and trace norm}

\begin{definition}
\label{def:app:trace_class}
An operator $A \in \cL(H)$ is \emph{trace class} if
\begin{equation}
\label{eq:app:trace_norm}
    \|A\|_1 \coloneqq \Tr|A| = \sum_{k=1}^\infty s_k(A) < \infty.
\end{equation}
The space of trace-class operators is $\cT{1}(H)$.
\end{definition}

\begin{proposition}[Trace]
\label{prop:app:trace}
For $A \in \cT{1}(H)$ and any orthonormal basis $\{e_k\}$, the series
\begin{equation}
\label{eq:app:trace}
    \Tr(A) \coloneqq \sum_{k=1}^\infty \langle e_k, A e_k\rangle
\end{equation}
converges absolutely, independently of the choice of basis.
\end{proposition}

\begin{proposition}
\label{prop:app:trace_class_properties}
$\cT{1}(H)$ satisfies:
\begin{enumerate}
\item[\textup{(i)}] $\cT{1}(H) \subset \cT{2}(H) \subset \cK(H)$,
    with $\|A\| \leq \|A\|_2 \leq \|A\|_1$.
\item[\textup{(ii)}] $(\cT{1}(H), \|\cdot\|_1)$ is a Banach space
    \textup{(}not a Hilbert space\textup{)}.
\item[\textup{(iii)}] $\cT{1}(H)$ is a two-sided ideal:
    $\|BA\|_1 \leq \|B\|\|A\|_1$.
\item[\textup{(iv)}] \emph{Duality:} the trace pairing
    $\langle A, X\rangle \coloneqq \Tr(A^*X)$
    identifies $\cT{1}(H)^* \cong \cL(H)$ isometrically.
\item[\textup{(v)}] Spectral decomposition: for self-adjoint
    $A \in \cTsa{1}(H)$, the series~\eqref{eq:app:spectral} holds
    with $\sum_k |\lambda_k| = \|A\|_1 < \infty$.
\end{enumerate}
\end{proposition}

\begin{remark}[Trace-class operators as projective tensor product]
\label{rem:app:trace_projective}
The identification
\begin{equation}
\label{eq:app:tc_projective}
    \cT{1}(H) \cong H \otimes_\pi H^*
\end{equation}
holds isometrically, where $\|\cdot\|_\pi$ is the projective tensor
norm on $H \otimes H^*$~\cite{Grothendieck1953,Ryan2002}. This places
the trace norm firmly within the framework of Chapter~\ref{chap:algtensor}:
\begin{equation}
\label{eq:app:norm_chain}
    \|\cdot\|_\varepsilon \leq \|\cdot\|_H \leq \|\cdot\|_1
    = \|\cdot\|_\pi \quad \text{on } H \otimes_a H^*,
\end{equation}
where $\|\cdot\|_H$ is the Hilbert tensor norm of $H \otimes_H H^*$
and $\|\cdot\|_\varepsilon$ is the injective norm. The trace norm is
the \emph{largest} reasonable crossnorm; the Hilbert--Schmidt norm
is a reasonable crossnorm strictly between $\|\cdot\|_\varepsilon$
and $\|\cdot\|_\pi$ when $H$ is infinite-dimensional. Both satisfy
condition~\condINJ\ of Definition~\ref{def:crossnorm}. The identification
$\cT{1}(H)^* \cong \cL(H)$ (Proposition~\ref{prop:app:trace_class_properties}(iv))
is precisely the duality $(\|\cdot\|_\pi)^* = \|\cdot\|_\varepsilon$
from Chapter~\ref{chap:algtensor}.
\end{remark}

\begin{remark}[The trace norm and quantum distinguishability]
For two density operators $\rho, \sigma \in \mathcal{D}(H)$ (defined
in Section~\ref{app:quantum:states}), the quantity
$\tfrac{1}{2}\|\rho - \sigma\|_1$ is the \emph{total variation distance}
between the two states: the maximum probability of distinguishing
$\rho$ from $\sigma$ by any quantum measurement. This operational
interpretation is why $\|\cdot\|_1$ is the natural norm for the
state space, in contrast to the Hilbert--Schmidt norm $\|\cdot\|_2$
which is computationally convenient but lacks a direct operational meaning.
\end{remark}

\subsection*{Partial trace}

Let $H = H_1 \otimes_H H_2$. For $A \in \cT{1}(H)$, the
\emph{partial trace over $H_2$} is the unique $\TrS{2}(A) \in \cT{1}(H_1)$
satisfying
\begin{equation}
\label{eq:app:partial_trace}
    \Tr_{H_1}\bigl(B\, \TrS{2}(A)\bigr)
    = \Tr_{H}\bigl((B \otimes I_2)\, A\bigr)
    \quad \forall\, B \in \cL(H_1).
\end{equation}
For any orthonormal basis $\{f_j\}$ of $H_2$ and
$\varphi_j = J(f_j) \in H_2^*$:
\begin{equation}
\label{eq:app:partial_trace_formula}
    \TrS{2}(A) = \sum_j (I_1 \otimes \varphi_j)\,
    A\, (I_1 \otimes J^{-1}(\varphi_j))^* \in \cT{1}(H_1),
\end{equation}
the series converging in $\|\cdot\|_1$.

\begin{proposition}[Partial trace and minimal subspaces]
\label{prop:app:partial_trace_min}
Let $v \in H_1 \otimes_H H_2$ with $\|v\|=1$ and let
$\rho_v = v \otimes J(v) \in \cT{1}(H)$ be the associated
rank-one operator. Then:
\begin{enumerate}
\item[\textup{(a)}] $\TrS{2}(\rho_v) \in \cT{1}(H_1)$ is positive
    with $\Tr(\TrS{2}(\rho_v)) = 1$ and
    \begin{equation*}
        \supp\bigl(\TrS{2}(\rho_v)\bigr)
        = \Umin{1}{v},
    \end{equation*}
    where $\Umin{1}{v}$ is the minimal subspace of $v$ in direction
    $\{1\}$ \textup{(}Definition~\ref{def:minimal_subspace}\textup{)}.
\item[\textup{(b)}] $\mathrm{rank}\bigl(\TrS{2}(\rho_v)\bigr)
    = r_1(v)$, the Tucker $1$-rank of $v$.
\item[\textup{(c)}] The non-zero eigenvalues of $\TrS{2}(\rho_v)$
    are $s_k(v)^2$, where $s_k(v)$ are the singular values of $v$
    viewed as a Hilbert--Schmidt operator $H_1 \to H_2$.
\end{enumerate}
\end{proposition}

\begin{proof}
Write $v = \sum_{k=1}^r s_k\, u_k \otimes_H w_k$ (Schmidt decomposition,
i.e., SVD of $v: H_1 \to H_2$). Then
$\rho_v = v \otimes J(v) = \sum_{j,k} s_j s_k
(u_j \otimes_H w_j) \otimes J(u_k \otimes_H w_k)$.
Using~\eqref{eq:app:partial_trace_formula} and $\langle w_l, w_j\rangle
= \delta_{lj}$:
\begin{equation*}
    \TrS{2}(\rho_v)
    = \sum_k s_k^2\, u_k \otimes J(u_k) \in \cT{1}(H_1).
\end{equation*}
This has $\supp = \spann\{u_k : s_k > 0\} = \Umin{1}{v}$,
rank $r_1(v)$, and eigenvalues $s_k^2$.
\end{proof}

%------------------------------------------------------------
\section{Density operators and quantum states}
\label{app:quantum:states}
%------------------------------------------------------------

\subsection*{The state space}

\begin{definition}
\label{def:app:density}
A \emph{density operator} on $H$ is an operator
$\rho \in \cTsa{1}(H)$ satisfying $\rho \geq 0$ and $\Tr(\rho) = 1$.
The \emph{state space} is
\begin{equation}
\label{eq:app:state_space}
    \Dstate{H}
    \coloneqq \bigl\{
    \rho \in \cTsa{1}(H) : \rho \geq 0,\;
    \Tr(\rho) = 1
    \bigr\}.
\end{equation}
\end{definition}

$\Dstate{H}$ is a convex subset of the Banach space $\cTsa{1}(H)$,
contained in the affine hyperplane $\{\Tr = 1\}$. It is
\emph{not} a manifold globally — it has rank singularities and
boundary phenomena — but each fixed-rank stratum is a smooth
Banach manifold; see Section~\ref{app:quantum:core} and Chapter~12.

\subsection*{Pure and mixed states}

\begin{definition}
\label{def:app:pure_mixed}
$\rho \in \Dstate{H}$ is \emph{pure} if $\mathrm{rank}(\rho) = 1$,
equivalently $\rho^2 = \rho$. Otherwise it is \emph{mixed}.
\end{definition}

Every pure state has the form $\rho = u \otimes J(u)$ for some unit
vector $u \in H$; two unit vectors $u$ and $e^{i\theta}u$ give the
same pure state. The set of pure states is in bijection with the
\emph{projective Hilbert space} $\mathbb{P}(H) = (H\setminus\{0\})/\CC^*$.

Every mixed state $\rho \in \Drank{r}{H}$ has a spectral decomposition
\begin{equation}
\label{eq:app:spectral_state}
    \rho = \sum_{k=1}^r \lambda_k\, v_k \otimes J(v_k),
\end{equation}
with $\{v_k\}$ orthonormal, $\lambda_k > 0$, $\sum_k \lambda_k = 1$.

\begin{remark}[The fundamental connection: minimal subspaces and quantum states]
\label{rem:app:states_minimal}
Let $v \in H$ be non-zero. The pure state
\begin{equation}
\label{eq:app:pure_state}
    \rho_v \coloneqq \frac{v \otimes J(v)}{\|v\|^2} \in \Dstate{H}
\end{equation}
has support $\supp(\rho_v) = \mathrm{Im}(\rho_v) = \spann\{v\}$,
which is the minimal subspace of $v$ in $H$.

In a multipartite system $H = \bigotimes_{j=1}^d H_j$, the partial
trace $\rho_\alpha \coloneqq \Tr_{\alpha^c}(\rho_v)$ satisfies, by
Proposition~\ref{prop:app:partial_trace_min}:
\begin{equation}
\label{eq:app:supp_minimal}
    \supp(\rho_\alpha) = \Umin{\alpha}{v},
    \qquad
    \mathrm{rank}(\rho_\alpha) = r_\alpha(v).
\end{equation}
This is the cornerstone of Chapter~12: the Tucker rank vector
$(r_\alpha(v))_{\alpha \in D}$ encodes the \emph{entanglement structure}
of the pure state $\rho_v$.
\end{remark}

\subsection*{The rank stratification}

\begin{definition}
\label{def:app:rank_strata}
The \emph{rank-$r$ stratum} is
$\Drank{r}{H} \coloneqq \{\rho \in \Dstate{H} : \mathrm{rank}(\rho) = r\}$.
\end{definition}

The decomposition $\Dstate{H} = \bigsqcup_{r \geq 1} \Drank{r}{H}
\cup \mathcal{D}_\infty(H)$ mirrors the disjoint decomposition
$\VD = \bigsqcup_\fr \Tucker{\fr}{\VD}$ of
Theorem~\ref{thm:tucker_decomposition_full}.

\begin{proposition}[Rank lower semicontinuity]
\label{prop:app:state_closure}
In the trace-norm topology, $\overline{\Drank{r}{H}}^{\|\cdot\|_1}
= \bigcup_{k=1}^r \Drank{k}{H}$: rank is weakly lower semicontinuous.
\end{proposition}

This is the quantum analogue of Theorem~\ref{thm:TBT_rank_lsc}.

%------------------------------------------------------------
\section{The positive core and support decomposition}
\label{app:quantum:core}
%------------------------------------------------------------

\subsection*{The positive core manifold}

Fix $S \in \Grass{r}{H}$ and let $V_S: \CC^r \to H$ be an isometry
with $\mathrm{Im}(V_S) = S$. Define
\begin{equation}
\label{eq:app:pos_core}
    \Dpos{S}
    \coloneqq \bigl\{
    \sigma \in \cTsa{1}(S) : \sigma > 0,\;
    \TrS{S}(\sigma) = 1
    \bigr\}
    \cong \Dpos{\CC^r},
\end{equation}
where the isomorphism is induced by $V_S$. Since $\dim S = r < \infty$,
$\cTsa{1}(S)$ is the space of self-adjoint $r \times r$ complex matrices
(isomorphic to $\RR^{r^2}$), and $\Dpos{S}$ is an open convex subset
of the hyperplane $\{\TrS{S} = 1\}$, a smooth manifold of real dimension
$r^2 - 1$.

\begin{remark}[Core as finite-dimensional Tucker fibre]
\label{rem:app:core_tucker}
$\Dpos{S}$ plays the role of the full-rank core space $\RR^{\times r_\alpha}_*$
from Chapter~\ref{chap:tucker_manifold}: it is the finite-dimensional
fibre of the bundle $\Drank{r}{H} \to \Grass{r}{H}$
(Theorem~3.8 of~\cite{AcevedoFalco2026} and Chapter~12 of this monograph).
\end{remark}

\subsection*{The support map and local coordinates}

The \emph{support map}
\begin{equation}
\label{eq:app:support_map}
    \pi: \Drank{r}{H} \to \Grass{r}{H},
    \qquad \pi(\rho) \coloneqq \supp(\rho) = \mathrm{Im}(\rho),
\end{equation}
is the quantum analogue of the rank map $\varrho_\fr$ of
Chapter~\ref{chap:tucker_manifold}.

\begin{proposition}[Local coordinates for $\Drank{r}{H}$]
\label{prop:app:local_coords}
Fix $\rho_0 \in \Drank{r}{H}$, $S_0 = \supp(\rho_0)$, and an isometry
$V_0: \CC^r \to H$ with $\mathrm{Im}(V_0) = S_0$. Let $P_0 = V_0 V_0^*
\in \cL(H)$ be the orthogonal projection onto $S_0$. For
$A \in \cL(\CC^r, S_0^\perp)$ sufficiently small in operator norm,
define
\begin{equation}
\label{eq:app:va}
    V_A \coloneqq (V_0 + \iota A)(I_{\CC^r} + A^*A)^{-1/2}
    \in \cL(\CC^r, H), \quad V_A^* V_A = I_{\CC^r},
\end{equation}
where $\iota: S_0^\perp \hookrightarrow H$ is the inclusion. Then:
\begin{enumerate}
\item[\textup{(i)}] The map
\begin{equation}
\label{eq:app:chart_states}
    \Phi_{S_0}: (A, \sigma)
    \longmapsto V_A \sigma V_A^*
    \in \Drank{r}{H}
\end{equation}
is a $\mathcal{C}^\infty$-diffeomorphism from a neighbourhood of
$(0, V_0^*\rho_0 V_0)$ in $\cL(\CC^r, S_0^\perp) \times \Dpos{\CC^r}$
onto a neighbourhood of $\rho_0$ in $\Drank{r}{H}$.
\item[\textup{(ii)}] The Grassmann variable $A \in \cL(\CC^r, S_0^\perp)$
encodes the motion of the support subspace
$S_0 \leadsto \mathrm{Im}(V_A)$, and the core variable
$\sigma \in \Dpos{\CC^r}$ encodes the positive degrees of freedom
within the support.
\end{enumerate}
\end{proposition}

This is the quantum specialisation of the Tucker manifold chart
(Lemma~\ref{lem:element}): the Grassmann variable $A$ plays the role
of $L_\alpha \in \cL(\Umin{\alpha}{\bv}, \Wmin{\alpha}{\bv})$, and
$\sigma$ plays the role of the core tensor $C^{(D)}$.

%------------------------------------------------------------
\section{Historical notes}
\label{app:quantum:history}
%------------------------------------------------------------

\paragraph{Mathematical foundations of quantum mechanics.}
The description of quantum states by density operators on Hilbert spaces
was introduced by von Neumann~\cite{vonNeumann1932}. His 1932 monograph
established the functional-analytic framework that remains standard today.
The use of mixed states (density operators of rank $> 1$) to describe
statistical mixtures appears in von Neumann's 1927 papers.

\paragraph{Schatten ideals.}
The systematic study of trace-class and Hilbert--Schmidt operators as
Banach ideals is due to Schatten~\cite{Schatten1950}, who introduced
the scale $\cT{p}(H)$ with norm $\|A\|_p = (\sum_k s_k(A)^p)^{1/p}$.
The duality $\cT{1}(H)^* \cong \cL(H)$ is the non-commutative
analogue of $(\ell^1)^* \cong \ell^\infty$.

\paragraph{Tensor norms and the trace class.}
The identification $\cT{1}(H) \cong H \otimes_\pi H^*$
(Remark~\ref{rem:app:trace_projective}) is due to
Grothendieck~\cite{Grothendieck1953}, who established it as part of
his classification of tensor norms. It places the trace norm firmly
within the framework of Chapter~\ref{chap:algtensor}: it is the
projective tensor norm on $H \otimes H^*$, the largest reasonable
crossnorm. The Hilbert--Schmidt identification $\cT{2}(H) \cong
H \otimes_H H^*$ (Theorem~\ref{thm:app:hs_tensor}) is the
non-commutative version of the $L^2$-tensor isomorphism $L^2(\Omega_1)
\otimes_H L^2(\Omega_2) \cong L^2(\Omega_1 \times \Omega_2)$
(Example~\ref{ex:Lp_tensor}).

\paragraph{The correct tensor product: $H \otimes H^*$ vs.\ $H \otimes H$.}
In the finite-dimensional literature, density matrices are often
written as elements of $H \otimes H$ via the identification
$H^* \cong H$ provided by the inner product. In infinite dimensions
this conflates two structurally distinct spaces: the dual $H^*$
(defined by the norm alone) and the conjugate $\overline{H}$ (defined
by the linear structure). Working with $H \otimes H^*$ throughout,
as we do in this appendix, keeps the tensor product framework of
Chapter~\ref{chap:algtensor} fully valid without any
infinite-dimensional caveats.

\paragraph{Minimal subspaces and partial traces.}
The identification of the support of the partial trace with the minimal
subspace (Proposition~\ref{prop:app:partial_trace_min} and
Remark~\ref{rem:app:states_minimal}) makes precise a connection implicit
in the quantum information literature since Schmidt (1907). The
systematic development as minimal subspaces in the context of tensor
Banach spaces is the contribution of~\cite{FH2012}, which forms the
algebraic foundation of Chapters~\ref{chap:minsubspaces} and~12.

\paragraph{The Banach manifold structure of $\Drank{r}{H}$.}
The local chart structure of Proposition~\ref{prop:app:local_coords}
is developed in full detail, with all proofs, in the companion
paper~\cite{AcevedoFalco2026}. The main results of that paper form
the content of Chapters~12 and~13 of this monograph.
