%============================================================
%  ch01_introduction.tex
%  Chapter 1: Introduction and Motivation
%
%  Tensor Formats in Banach Spaces
%  Antonio Falcó, CEU Universities
%
%  Estimated length: ~25 pp.
%============================================================

\chapter{Introduction and Motivation}
\label{chap:intro}

%------------------------------------------------------------
\section{The curse of dimensionality}
\label{sec:intro:curse}
%------------------------------------------------------------

Many of the central problems in modern mathematics and its applications
involve functions, operators, or data defined on high-dimensional
product domains. A $d$-variate function $f: I_1 \times \cdots \times
I_d \to \RR$ defined on a product of $d$ intervals, for example, lives
in a space whose dimension grows exponentially with $d$: if each
factor $I_\alpha$ is discretised with $n$ grid points, the natural
approximation space has dimension $n^d$. For $d = 100$ and $n = 10$
this is $10^{100}$, a number that no computer will ever handle directly.
This exponential explosion of complexity was famously termed by Bellman
the \emph{curse of dimensionality}.

High-dimensional problems of this kind arise throughout science and
engineering:
\begin{itemize}
\item \emph{Quantum chemistry:} the electronic Schrödinger equation
    for an $N$-electron molecule lives in $\RR^{3N}$; realistic
    molecules have $N$ in the hundreds.
\item \emph{Uncertainty quantification:} the solution of a PDE with
    $d$ uncertain parameters is a function on $\RR^d \times \Omega$,
    where $\Omega$ is the spatial domain.
\item \emph{Stochastic control and finance:} value functions in optimal
    control problems typically satisfy PDEs in $\RR^d$ for large $d$.
\item \emph{Machine learning:} the approximation of unknown functions
    from data in $\RR^d$ is the fundamental task of supervised learning.
\end{itemize}

The common mathematical structure underlying all these problems is that
the object of interest — a function, a quantum state, a probability
density — is an element of a \emph{tensor product space}
${}_a\bigotimes_{\alpha \in D} V_\alpha$, where $D = \{1,\ldots,d\}$
and $V_\alpha$ is a function space on the $\alpha$-th variable.
The curse of dimensionality is the statement that the dimension of this
space grows exponentially with $\card{D} = d$.

The strategy for circumventing the curse is to restrict attention to
\emph{structured low-rank tensors}: elements of $\VD$ that can be
represented using far fewer parameters than $n^d$. The systematic study
of such representations — their algebraic structure, their topological
properties in function spaces, and their geometry as differentiable
manifolds — is the subject of this monograph.

%------------------------------------------------------------
\section{Tensor formats and the role of subspaces}
\label{sec:intro:formats}
%------------------------------------------------------------

\subsection*{From matrices to tensors}

The simplest non-trivial case is $d = 2$: here $\VD = V_1 \atens V_2$,
and a tensor $\bv \in \VD$ corresponds to a linear map (a ``matrix'')
from $V_1^*$ to $V_2$. The \emph{rank} of $\bv$ — the minimal $r$
such that $\bv = \sum_{\nu=1}^r u_\nu \otimes w_\nu$ — determines the
complexity of the representation. In finite dimensions this is the rank
of the associated matrix; the Eckart--Young theorem gives the best
rank-$r$ approximation in the spectral or Frobenius norm.

The higher-order analogue of rank, however, is not straightforward.
For $d \geq 3$ there is no single notion of ``tensor rank'' with all
the good properties of matrix rank. Instead, several notions of rank
have proven useful, each corresponding to a different way of
decomposing the factors of $\VD$.

\subsection*{Tucker format}

The \emph{Tucker format}~\cite{Tucker1966} represents a tensor $\bv$
as
\begin{equation*}
    \bv = \sum_{\substack{1 \leq i_\alpha \leq r_\alpha \\ \alpha \in D}}
    C^{(D)}_{(i_\alpha)_{\alpha \in D}}
    \bigotimes_{\alpha \in D} u^{(\alpha)}_{i_\alpha},
\end{equation*}
where $u^{(\alpha)}_{i_\alpha} \in V_\alpha$ and
$C^{(D)} \in \RR^{r_1 \times \cdots \times r_d}$ is the \emph{core tensor}.
The storage cost is $\sum_\alpha r_\alpha n_\alpha + \prod_\alpha r_\alpha$,
which is linear in $d$ when the ranks $r_\alpha$ are bounded. The Tucker
format is equivalent to the \emph{multilinear SVD} (HOSVD) of De Lathauwer,
De Moor, and Vandewalle~\cite{DeLathauwer2000}.

The correct functional-analytic description of the Tucker format is
through \emph{minimal subspaces}: the minimal subspace
$\Umin{\alpha}{\bv} \subset V_\alpha$ of $\bv$ is the smallest
subspace such that $\bv \in \Umin{\alpha}{\bv} \atens V_{[\alpha]}$
(Chapter~\ref{chap:minsubspaces}). The Tucker rank of $\bv$ is the
tuple $(r_\alpha)_{\alpha \in D}$ where $r_\alpha = \dim \Umin{\alpha}{\bv}$.

\subsection*{Hierarchical Tucker and Tensor Train formats}

The Tucker format still suffers from the curse of dimensionality through
its core tensor: $\prod_\alpha r_\alpha$ grows exponentially with $d$.
The \emph{Hierarchical Tucker} (HT) format~\cite{HaKuehn2009} and the
\emph{Tensor Train} (TT) format~\cite{Oseledets2011} resolve this by
organising the dimensions into a \emph{hierarchy}, encoded by a rooted tree
$\TD$. Each format restricts the tensor to satisfy rank constraints at
every level of the tree, reducing the total storage to $O(d r^3)$ (HT)
or $O(d r^2 n)$ (TT), where $r$ is the maximal rank.

\subsection*{Tree-based tensor formats}

This monograph studies all these formats within a single unified framework:
\emph{tree-based Tucker formats}, introduced in~\cite{FHN2021}. Given a
dimension partition tree $\TD$ over $D$ and a rank tuple $\fr$, the set
\begin{equation*}
    \FTv = \bigcap_{k=1}^{\dep{\TD}}
    \Tucker{\fr_k}{\VD, \PkT{k}{\TD}}
\end{equation*}
of tensors with fixed tree-based rank $\fr$ is an intersection of Tucker
sets at each level of the tree. The Tucker format is the special case
$\TD = \TDtuck$ (depth~1); HT and TT correspond to binary trees of
depth $\lceil \log_2 d \rceil$ and $d-1$ respectively.

%------------------------------------------------------------
\section{Three fundamental questions}
\label{sec:intro:questions}
%------------------------------------------------------------

The results of this monograph are organised around three fundamental
mathematical questions, each corresponding to one part of the book.

\paragraph{Question 1 (Chapter~\ref{chap:minsubspaces}).}
\emph{Does every tensor $\bv$ in an infinite-dimensional Banach tensor
space $\VDnorm$ have a best approximation from the set
$\FTleq{\fr}{\VD}{\TD}$?}

For convex sets in reflexive Banach spaces, best approximations always
exist. The set $\FTleq{\fr}{\VD}{\TD}$ is not convex (it is not even a
subspace), so the existence of best approximations is non-trivial.
The answer, established in Chapters~\ref{chap:minsubspaces}
and~\ref{chap:topTBT}, is \emph{yes}: under a natural condition on the
norm (condition \condINJ, which holds for all $L^p$ and Sobolev spaces),
$\FTleq{\fr}{\VD}{\TD}$ is weakly closed in $\VDnorm$, and reflexivity
gives the best approximation.

\paragraph{Question 2 (Chapters~\ref{chap:tucker_manifold}
and~\ref{chap:TB_manifold}).}
\emph{Does the set $\Tucker{\fr}{\VD}$ (or $\FTv$) of tensors with
fixed rank carry the structure of a differentiable manifold, even in
infinite dimensions?}

In finite dimensions, $\Tucker{\fr}{\VD}$ is known to be a smooth
manifold (a quotient of a matrix group). In infinite dimensions, the
standard quotient construction fails because the stabiliser group is
infinite-dimensional and the quotient topology may not be well-behaved.
The answer, established in~\cite{FHN2019} for Tucker formats and
in~\cite{FHN2023} for tree-based formats, is \emph{yes}: an explicit
atlas with $\mathcal{C}^\infty$ transition maps can be constructed
using the Grassmann--Banach manifold structure of the minimal subspaces.

\paragraph{Question 3 (Chapter~\ref{chap:DF}).}
\emph{Can the Dirac--Frenkel variational principle — the projection
of the vector field onto the tangent space of the manifold — be
extended from Hilbert to Banach tensor spaces?}

The answer is \emph{yes}, provided the tangent space $\ZD{\bv}$ is a
\emph{split subspace} of $\VDnorm$ (i.e., has a bounded projection
onto it). The immersion theorems of Chapters~\ref{chap:tucker_manifold}
and~\ref{chap:TB_manifold} — the deepest results of the monograph —
establish exactly this.

%------------------------------------------------------------
\section{The four source papers}
\label{sec:intro:papers}
%------------------------------------------------------------

This monograph synthesises and expands four research papers written
by the author and his collaborators Wolfgang Hackbusch and Anthony Nouy:

\medskip

\noindent\textbf{[FH2012]} \emph{On Minimal Subspaces in Tensor
Representations}~\cite{FH2012}. This paper introduces minimal subspaces
as the fundamental algebraic object in tensor representations, proves
their existence and uniqueness in general vector spaces, and establishes
the existence of best Tucker approximations in reflexive Banach tensor
spaces. It constitutes Part~I (Chapters~\ref{chap:algtensor}--\ref{chap:minsubspaces})
of this monograph.

\medskip

\noindent\textbf{[FHN2019]} \emph{On the Dirac--Frenkel Variational
Principle on Tensor Banach Spaces}~\cite{FHN2019}. This paper
constructs the $\mathcal{C}^\infty$-Banach manifold structure of
$\Tucker{\fr}{\VD}$ via explicit local charts, proves that the manifold
is immersed in $\VDnorm$ under condition \condINJ, and extends the
Dirac--Frenkel variational principle to Banach tensor spaces. It forms
the basis of Chapters~\ref{chap:tucker_manifold}
and~\ref{chap:DF}.

\medskip

\noindent\textbf{[FHN2021]} \emph{Tree-based Tensor
Formats}~\cite{FHN2021}. This paper introduces the general framework
of dimension partition trees, proves the existence of best tree-based
approximations under conditions weaker than~\cite{FH2012}, and
characterises tensors of fixed tree-based rank by means of transfer
tensors. It underlies Chapters~\ref{chap:trees}--\ref{chap:topTBT}.

\medskip

\noindent\textbf{[FHN2023]} \emph{Geometry of Tree-based Tensor
Formats in Tensor Banach Spaces}~\cite{FHN2023}. This paper provides
the geometric description of $\FTv$ as an intersection of Tucker
manifolds, proves that it is a $\mathcal{C}^\infty$-Banach manifold
compatible with the Tucker structure at each level of the tree, and
extends the Dirac--Frenkel principle to tree-based formats.
It forms the basis of Chapter~\ref{chap:TB_manifold}.

%------------------------------------------------------------
\section{What this monograph adds}
\label{sec:intro:added}
%------------------------------------------------------------

Beyond collecting and unifying the results of the four source papers,
this monograph contributes the following:

\begin{enumerate}
\item \textbf{Unified notation.} The four papers were written over a
    decade and use slightly different notational conventions. This
    monograph adopts a single coherent system (detailed in the
    Notation chapter), in which the Tucker format is the special
    case $\TD = \TDtuck$ of the tree-based format, minimal subspaces
    are always denoted $\Umin{\alpha}{\bv}$, and the model Banach
    space of the Tucker manifold is always $\Emodel{D}$.

\item \textbf{Expanded proofs.} Several results in the papers are stated
    with proof sketches or references to earlier work. This monograph
    provides complete proofs of the central theorems, including the
    full proof of the local chart system (Lemma~\ref{lem:element}),
    the $\mathcal{C}^\infty$ transition maps
    (Lemma~\ref{lem:premanifold}), the tangent space identification
    (Proposition~\ref{prop:tangent_space}), and the immersion theorems
    (Theorems~\ref{thm:tucker_immersion} and~\ref{thm:TBT_immersion}).

\item \textbf{Self-contained background.} Chapters~\ref{chap:banach}
    and~\ref{chap:algtensor} provide the Banach space and algebraic
    tensor product prerequisites in the form and notation used
    throughout, so that the monograph is accessible without consulting
    external references for the background.

\item \textbf{Historical context and comparison.} Each chapter closes
    with historical notes that place the results in context and compare
    them with alternative approaches in the literature.

\item \textbf{Open problems.} Chapter~\ref{chap:open} collects ten
    concrete open problems arising naturally from the theory, ranging
    from the global topology of tensor manifolds to learning theory on
    tree-based manifolds.
\end{enumerate}

%------------------------------------------------------------
\section{Overview of the main results}
\label{sec:intro:overview}
%------------------------------------------------------------

We summarise the principal theorems of the monograph, grouped by theme.

\paragraph{Existence of best approximations
(Chapters~\ref{chap:minsubspaces},~\ref{chap:topTBT}).}

\begin{itemize}
\item \emph{Theorem~\ref{thm:best_approx_tucker}} [FH2012]: In a
    reflexive Banach tensor space $\VDnorm$ satisfying \condINJ,
    every $\bu \in \VDnorm$ has a best Tucker approximation of any
    prescribed rank.
\item \emph{Theorem~\ref{thm:TBT_best_approx}} [FHN2021]: The same
    holds for tree-based formats under the weaker local condition
    \condINJ[$T_D$].
\end{itemize}

\paragraph{Manifold structure
(Chapters~\ref{chap:tucker_manifold},~\ref{chap:TB_manifold}).}

\begin{itemize}
\item \emph{Theorem~\ref{thm:Tucker_Banach_Manifold}} [FHN2019]:
    If the tensor product map is continuous, $\Tucker{\fr}{\VD}$
    is a $\mathcal{C}^\infty$-Banach manifold modelled on
    $\Emodel{D} = \prod_\alpha \cL(\Umin{\alpha}{\bv},
    \Wmin{\alpha}{\bv}) \times \RR^{\times r_\alpha}$.
    It is a fibre bundle over $\prod_\alpha \Grass{r_\alpha}{V_\alpha}$
    with fibre $\RR^{\times r_\alpha}_*$.
\item \emph{Theorem~\ref{thm:TBT_Banach_Manifold}} [FHN2023]:
    Under \condINJ\ at each level, $\FTv$ is a $\mathcal{C}^\infty$-Banach
    manifold, with an atlas compatible with the Tucker atlases at
    each level of $\TD$.
\end{itemize}

\paragraph{Immersion theorems
(Chapters~\ref{chap:tucker_manifold},~\ref{chap:TB_manifold}).}

\begin{itemize}
\item \emph{Theorem~\ref{thm:tucker_immersion}} [FHN2019]:
    Under \condINJ, the tangent subspace $\ZD{\bv}$ is a split
    subspace of $\VDnorm$, and $\Tucker{\fr}{\VD}$ is an immersed
    submanifold of $\VDnorm$.
\item \emph{Theorem~\ref{thm:TBT_immersion}} [FHN2023]:
    The same holds for $\FTv$, with tangent space
    $\bigcap_k \ZD[\PkT{k}{\TD}]{\bv}$.
\end{itemize}

\paragraph{Dirac--Frenkel variational principle (Chapter~\ref{chap:DF}).}

\begin{itemize}
\item \emph{Theorem~\ref{thm:DF_tucker}} [FHN2019]: In a reflexive,
    strictly convex $\VDnorm$ satisfying \condINJ, the Dirac--Frenkel
    ODE on $\Tucker{\fr}{\VD}$ has local solutions, and they are
    characterised by the metric projection onto $\ZD{\bv_r(t)}$.
\item \emph{Theorem~\ref{thm:DF_treebased}} [FHN2023]: The same
    holds on $\FTv$.
\end{itemize}

%------------------------------------------------------------
\section{Guide to the reader}
\label{sec:intro:guide}
%------------------------------------------------------------

The monograph is organised into four parts.

\textbf{Part~\ref{part:foundations}: Algebraic and Topological Foundations}
(Chapters~\ref{chap:banach}--\ref{chap:minsubspaces}).
Chapter~\ref{chap:banach} collects the necessary background from Banach
space theory and infinite-dimensional differential geometry.
Chapter~\ref{chap:algtensor} develops the algebraic theory of tensor
products. Chapter~\ref{chap:minsubspaces} is the algebraic core of the
monograph: it introduces minimal subspaces, proves their fundamental
properties, and establishes the existence of best approximations.
\emph{Readers familiar with Banach manifolds may skip
Chapter~\ref{chap:banach} and refer back as needed.}

\textbf{Part~\ref{part:tree}: Tree-Based Tensor Formats}
(Chapters~\ref{chap:trees}--\ref{chap:topTBT}).
Chapter~\ref{chap:trees} introduces dimension partition trees and
proves the intersection representation $\FTv = \bigcap_k \Tucker{\fr_k}{}$.
Chapter~\ref{chap:algTBT} develops the algebraic theory of tree-based
formats, culminating in the transfer tensor representation
(Theorem~\ref{thm:TBT_characterisation}).
Chapter~\ref{chap:topTBT} treats the topological theory: the
$T_D$-continuity condition, the identification theorem, and the
existence of best tree-based approximations.
\emph{Readers primarily interested in the Tucker format may focus on
Chapters~\ref{chap:minsubspaces} and then jump to
Part~\ref{part:geometry}.}

\textbf{Part~\ref{part:geometry}: Differential Geometry of Tensor Formats}
(Chapters~\ref{chap:tucker_manifold}--\ref{chap:TB_manifold}).
These two chapters contain the geometrically deepest results of the
monograph. Chapter~\ref{chap:tucker_manifold} constructs the
$\mathcal{C}^\infty$ Tucker manifold and proves its immersion in
$\VDnorm$. Chapter~\ref{chap:TB_manifold} extends these results to
tree-based formats via the Laplacian map and the intersection of Tucker
atlases. \emph{The key technical result is the immersion theorem
(Theorems~\ref{thm:tucker_immersion} and~\ref{thm:TBT_immersion}),
which relies on Lemmas~\ref{lem:two_faces} and~\ref{lem:direct_sum_G}.}

\textbf{Part~\ref{part:variational}: Variational Principles and Applications}
(Chapters~\ref{chap:DF}--\ref{chap:open}).
Chapter~\ref{chap:DF} applies the geometric results to the
Dirac--Frenkel variational principle, establishing the existence and
characterisation of solutions to the reduced ODE on tensor manifolds.
Chapter~\ref{chap:open} collects ten open problems.

The three appendices (Appendices~\ref{app:manifolds}--\ref{app:notation_index})
provide quick references for Banach manifold theory, functional analysis,
and the notation.

\paragraph{Prerequisites.} The reader is assumed to be familiar with
functional analysis at the level of a standard graduate course (Banach
spaces, weak convergence, Hilbert spaces) and with differential
manifolds at the level of a first course using smooth finite-dimensional
manifolds. The infinite-dimensional theory (Banach manifolds,
Grassmannians) is developed from scratch in Chapter~\ref{chap:banach}.

\paragraph{Notation.} A complete list of symbols is given in the
Notation chapter following this Introduction. All macros used in the
text are defined in the file \texttt{notation.tex}.

%------------------------------------------------------------
\section{Historical context}
\label{sec:intro:history}
%------------------------------------------------------------

The mathematical study of tensor approximation has roots in several
independent traditions.

In \emph{multilinear algebra}, Hitchcock~\cite{Hitchcock1927} introduced
the notion of tensor rank in 1927. Tucker~\cite{Tucker1966} introduced
the Tucker format in psychometrics in 1966. The HOSVD was introduced by
De Lathauwer, De Moor, and Vandewalle~\cite{DeLathauwer2000} in 2000.

In \emph{quantum physics}, the matrix product state (MPS) representation
appeared in Vidal's work~\cite{Vidal2003} on quantum simulation and was
independently proposed as the TT decomposition by Oseledets and
Tyrtyshnikov~\cite{Oseledets2011}. The DMRG algorithm of White, which
is equivalent to optimisation over MPS/TT manifolds, dates from 1992.

In \emph{numerical analysis}, the HT format was introduced by Hackbusch
and Kühn~\cite{HaKuehn2009} with the explicit goal of controlling storage
costs for high-dimensional PDEs. The existence of best approximations
in infinite-dimensional tensor spaces was first established in the
present framework by Falcó and Hackbusch~\cite{FH2012}.

The \emph{differential-geometric} treatment of tensor manifolds began
with Koch and Lubich~\cite{KochLubich2007} in the finite-dimensional
Hilbert space setting, motivated by the MCTDH method. The extension to
infinite-dimensional Banach spaces in~\cite{FHN2019,FHN2023} required
new ideas — notably the explicit chart construction via Grassmann charts
and the immersion theorem — that form the core of this monograph.

