%============================================================
%  ch02_banach.tex
%  Chapter 2: Banach Space Prerequisites
%
%  Tensor Formats in Banach Spaces
%  Antonio Falcó, CEU Universities
%
%  Estimated length: ~30 pp.
%============================================================

\chapter{Banach Space Prerequisites}
\label{chap:banach}

This chapter collects the functional-analytic and differential-geometric
background required throughout the monograph. The reader familiar with
Banach space theory and infinite-dimensional differential geometry may
proceed directly to Chapter~\ref{chap:algtensor}, referring back as needed.
All vector spaces are over the field $\KK$, which is either $\RR$ or $\CC$;
unless stated otherwise, $\KK = \RR$.

%------------------------------------------------------------
\section{Normed spaces and Banach spaces}
\label{sec:banach:basics}
%------------------------------------------------------------

We begin by fixing notation and recalling the central notions of normed and
Banach space theory that will be used throughout.

\subsection*{Basic definitions}

Let $X$ be a vector space over $\KK$. A \emph{norm} on $X$ is a map
$\|\cdot\|: X \to [0,\infty)$ satisfying:
\begin{enumerate}
\item[(N1)] $\|x\| = 0$ if and only if $x = 0$,
\item[(N2)] $\|\lambda x\| = |\lambda|\,\|x\|$ for all $\lambda \in \KK$,
    $x \in X$,
\item[(N3)] $\|x + y\| \leq \|x\| + \|y\|$ for all $x, y \in X$.
\end{enumerate}
The pair $(X, \|\cdot\|)$ is called a \emph{normed space}. A sequence
$(x_n)_{n \in \NN}$ in $X$ is a \emph{Cauchy sequence} if
$\|x_n - x_m\| \to 0$ as $n, m \to \infty$. A normed space in which every
Cauchy sequence converges is called a \emph{Banach space}.

Two norms $\|\cdot\|$ and $\|\cdot\|'$ on $X$ are \emph{equivalent} if
there exist constants $c, C > 0$ such that
$c\|x\| \leq \|x\|' \leq C\|x\|$ for all $x \in X$.
Equivalent norms define the same topology, and hence the same Cauchy
sequences, so they produce the same Banach space structure.

\begin{example}
\label{ex:banach:lp}
Let $I \subset \RR$ be a measurable set and $1 \leq p < \infty$. The
Lebesgue space $L^p(I)$, equipped with the norm
$\|f\|_{L^p} = \bigl(\int_I |f(x)|^p\,dx\bigr)^{1/p}$,
is a Banach space. For $p = \infty$ one uses the essential supremum norm.
The case $p = 2$ gives a Hilbert space with inner product
$\langle f, g\rangle = \int_I f(x)\overline{g(x)}\,dx$.
\end{example}

\begin{example}
\label{ex:banach:sobolev}
For a bounded open set $\Omega \subset \RR^n$, $1 \leq p < \infty$, and
$k \in \NNz$, the Sobolev space $W^{k,p}(\Omega)$ consists of all functions
$f \in L^p(\Omega)$ whose distributional derivatives $D^\alpha f$ of order
$|\alpha| \leq k$ belong to $L^p(\Omega)$. Equipped with the norm
\begin{equation*}
    \|f\|_{W^{k,p}} = \Bigl(\sum_{|\alpha| \leq k}
        \|D^\alpha f\|_{L^p}^p\Bigr)^{1/p},
\end{equation*}
it is a Banach space. For $p=2$ one writes $H^k(\Omega) = W^{k,2}(\Omega)$,
which is a Hilbert space. These spaces appear as natural component spaces in
high-dimensional tensor problems; see Example~\ref{ex:min:sobolev} and
Section~\ref{sec:topTBT:bestapprox}.
\end{example}

\subsection*{Continuous linear maps}

Let $X$ and $Y$ be normed spaces. A linear map $A: X \to Y$ is
\emph{continuous} if and only if it is \emph{bounded}, meaning
\begin{equation*}
    \opnorm{A}{Y}{X} \coloneqq \sup_{x \neq 0} \frac{\|Ax\|_Y}{\|x\|_X}
    = \sup_{\|x\|_X \leq 1} \|Ax\|_Y < \infty.
\end{equation*}
The quantity $\opnorm{A}{Y}{X}$ is the \emph{operator norm} of $A$.
The space of all continuous linear maps from $X$ to $Y$, denoted
$\cL(X, Y)$, is itself a normed space under the operator norm; it is a
Banach space whenever $Y$ is complete. We write $\cL(X) \coloneqq \cL(X,X)$.
The set of bounded linear automorphisms (bounded invertible maps with
bounded inverse) is
\begin{equation*}
    \GL{X} \coloneqq \{A \in \cL(X) : A \text{ invertible}\}.
\end{equation*}
This is an open subset of $\cL(X)$, and hence a Banach manifold modelled
on $\cL(X)$; see Example~\ref{ex:manifold:GL}.

\subsection*{Dual spaces and reflexivity}

The \emph{topological dual} of a normed space $(X, \|\cdot\|)$ is
$\dual{X} \coloneqq \cL(X, \KK)$, equipped with the dual norm
\begin{equation}
\label{eq:dualnorm}
    \|\varphi\|_{\dual{X}} \coloneqq \sup_{\|x\| \leq 1} |\varphi(x)|.
\end{equation}
The dual $\dual{X}$ is always a Banach space, regardless of whether $X$
is complete. The \emph{bidual} $X^{**} \coloneqq (\dual{X})^*$ contains
$X$ as a canonically embedded subspace via the isometric embedding
$\iota: X \to X^{**}$, $\iota(x)(\varphi) = \varphi(x)$. When $\iota$ is
surjective, $X$ is called \emph{reflexive}. All $L^p$ spaces with
$1 < p < \infty$ are reflexive; $L^1$ and $L^\infty$ are not.

Reflexivity plays an important role in the existence of best approximations.
Recall that a closed set $C$ in a normed space $X$ is \emph{proximinal} if
every point of $X$ has at least one nearest point in $C$. In a reflexive
Banach space, every closed convex set is proximinal
(see, e.g., \cite[p.\,61]{Hackbusch2019}). This fact will be used
extensively in Chapter~\ref{chap:minsubspaces} to prove the existence
of best low-rank tensor approximations.

\subsection*{Convexity and smoothness}

Two geometric properties of the norm are of particular importance for
the Dirac--Frenkel variational principle in Chapter~\ref{chap:DF}.

\begin{definition}
\label{def:strictly_convex}
A normed space $(X, \|\cdot\|)$ is \emph{strictly convex} (or
\emph{rotund}) if $\|x + y\|/2 < 1$ for all $x, y \in X$ with
$\|x\| = \|y\| = 1$ and $x \neq y$.
\end{definition}

\begin{definition}
\label{def:uniformly_convex}
A normed space $(X, \|\cdot\|)$ is \emph{uniformly convex} if for every
$\varepsilon > 0$ there exists $\delta > 0$ such that
$\|x + y\|/2 \leq 1 - \delta$ whenever $\|x\|, \|y\| \leq 1$ and
$\|x - y\| \geq \varepsilon$.
\end{definition}

Every uniformly convex Banach space is reflexive and strictly convex
(Clarkson's theorem). In particular, $L^p$ spaces with $1 < p < \infty$
are uniformly convex. Strict convexity guarantees uniqueness of metric
projections onto closed convex sets, a property needed for the
well-posedness of the Dirac--Frenkel equation.

\begin{definition}
\label{def:smooth}
A Banach space $X$ is \emph{smooth} if the limit
$\lim_{t \to 0} \bigl(\|x + ty\| - \|x\|\bigr)/t$
exists for all $x, y \in X$ with $\|x\| = 1$.
\end{definition}

Smoothness is the dual property to strict convexity in the following
sense: $X$ is smooth if and only if $\dual{X}$ is strictly convex, and
$X$ is strictly convex if and only if $\dual{X}$ is smooth.

%------------------------------------------------------------
\section{Complemented subspaces and projections}
\label{sec:banach:complemented}
%------------------------------------------------------------

The central geometric notion underlying all manifold structures in this
monograph is that of a complemented (or split) subspace. Unlike the
Hilbert space setting, complementability is a non-trivial condition in
general Banach spaces.

\subsection*{Projections}

A bounded linear map $P \in \cL(X)$ is a \emph{projection} if $P^2 = P$.
In that case $\mathrm{Im}\,P$ is a closed subspace, $\mathrm{Ker}\,P$ is
closed, and $X = \mathrm{Im}\,P \oplus \mathrm{Ker}\,P$ (as a direct sum
of closed subspaces). The complementary projection is $\Id{X} - P$.

\begin{definition}
\label{def:split_subspace}
Let $X$ be a Banach space. A closed subspace $U \subset X$ is a
\emph{split subspace} (or \emph{complemented subspace}) of $X$ if
there exists a closed subspace $W \subset X$ such that
$X = U \oplus W$.
The pair $(U, W)$ is then called a \emph{pair of complementary
subspaces}, and $W$ is a \emph{(topological) complement} of $U$ in $X$.
We write $\Proj{U}{W}$ for the unique continuous projection from $X$
onto $U$ along $W$.
\end{definition}

The following proposition characterises split subspaces in terms of
projections and provides three equivalent conditions that will each be
used at different points of the text.

\begin{proposition}[{\cite[Proposition~4.2]{FHN2015}}]
\label{prop:split_equiv}
Let $X$ be a Banach space and $U \subset X$ a closed subspace. The
following conditions are equivalent:
\begin{enumerate}
\item[\textup{(a)}] $U$ is a split subspace of $X$.
\item[\textup{(b)}] There exists $P \in \cL(X)$ with $P^2 = P$ and
    $\mathrm{Im}\,P = U$.
\item[\textup{(c)}] There exists $Q \in \cL(X)$ with $Q^2 = Q$ and
    $\mathrm{Ker}\,Q = U$.
\end{enumerate}
\end{proposition}

In condition~(b), the projection $P$ is none other than $\Proj{U}{W}$
for any choice of closed complement $W$ of $U$; in~(c), one has $Q =
\Proj{W}{U}$.

\begin{proposition}[{\cite[Theorem~4.5]{FHN2015}}]
\label{prop:finitedim_split}
Let $X$ be a Banach space. Every finite-dimensional subspace of $X$ is
a split subspace.
\end{proposition}

In a Hilbert space, every closed subspace is complemented (the orthogonal
complement provides the split). This is no longer true in general Banach
spaces: by a classical result of Lindenstrauss and Tzafriri, a Banach
space in which every closed subspace is complemented is necessarily
isomorphic to a Hilbert space. The failure of complementability in
non-Hilbert spaces is precisely what makes the immersion theorems of
Chapter~\ref{chap:tucker_manifold} non-trivial.

\subsection*{The nilpotent algebra associated to a splitting}

The following result, central to the construction of local charts in
Chapters~\ref{chap:tucker_manifold} and~\ref{chap:TB_manifold},
identifies a particularly well-behaved sub-algebra of $\cL(X)$
associated to any splitting $X = U \oplus W$.

\begin{proposition}
\label{prop:nilpotent_algebra}
Let $X$ be a Banach space and $U, W \in \GrassB{X}$ with $X = U \oplus W$.
Define
\begin{equation*}
    \cL_{(U,W)}(X) \coloneqq
    \bigl\{
        \Proj{W}{U} \circ S \circ \Proj{U}{W} : S \in \cL(X)
    \bigr\}
    \cong \cL(U, W).
\end{equation*}
Then:
\begin{enumerate}
\item[\textup{(i)}] The natural map
    $\cL(U, W) \to \cL_{(U,W)}(X)$,
    $L \mapsto \Proj{W}{U} \circ L \circ \Proj{U}{W}$,
    is an isometric isomorphism.
\item[\textup{(ii)}] $\cL_{(U,W)}(X)$ is a nilpotent sub-algebra of
    $\cL(X)$: for every $L, L' \in \cL_{(U,W)}(X)$ one has $L \circ L' = 0$.
\item[\textup{(iii)}] As a consequence, for every
    $L \in \cL_{(U,W)}(X)$:
    \begin{equation}
    \label{eq:exp_nilpotent}
        \exp(L) \coloneqq \sum_{n=0}^{\infty} \frac{L^n}{n!}
        = \Id{X} + L,
        \qquad
        \exp(-L) = \Id{X} - L = (\Id{X} + L)^{-1}.
    \end{equation}
\end{enumerate}
\end{proposition}

\begin{proof}
For the isometry in~(i), note that
$\|\Proj{W}{U} \circ L \circ \Proj{U}{W}\|_{X \leftarrow X}
= \|L\|_{W \leftarrow U}$ by the definition of the operator norm.
For~(ii): if $L = \Proj{W}{U} \circ S \circ \Proj{U}{W}$ and
$L' = \Proj{W}{U} \circ S' \circ \Proj{U}{W}$, then
\begin{equation*}
    L \circ L' =
    \Proj{W}{U} \circ S \circ (\Proj{U}{W} \circ \Proj{W}{U})
    \circ S' \circ \Proj{U}{W} = 0,
\end{equation*}
since $\Proj{U}{W} \circ \Proj{W}{U} = 0$. Statement~(iii) follows
immediately from~(ii) and the series definition of the exponential.
\end{proof}

The identity $\exp(L) = \Id{X} + L$ means that the exponential map on
the nilpotent algebra $\cL_{(U,W)}(X)$ is a polynomial of degree one,
and in particular a global analytic diffeomorphism. This is the key
that makes the local charts of the Grassmann--Banach manifold and the
Tucker manifold globally well-defined; see
Section~\ref{sec:banach:grassmann}.

\subsection*{Common complements and chart compatibility}

The following lemma, which underlies the compatibility of overlapping
charts in the Grassmann--Banach manifold, characterises when two split
subspaces share a common complement.

\begin{lemma}[{\cite[Lemma~2.1]{DixmierDouady1963}}]
\label{lem:common_complement}
Let $X$ be a Banach space and let $U, U', W \in \GrassB{X}$ with
$X = U \oplus W = U' \oplus W$. The following conditions are equivalent:
\begin{enumerate}
\item[\textup{(a)}] $U$ and $U'$ have $W$ as a common complement, i.e.,
    $X = U \oplus W = U' \oplus W$.
\item[\textup{(b)}] The restriction $\Proj{U}{W}\big|_{U'} : U' \to U$
    is a linear isomorphism (with bounded inverse).
\end{enumerate}
Moreover, in this case
$\bigl(\Proj{U}{W}\big|_{U'}\bigr)^{-1} = \Proj{U'}{W}\big|_U$.
\end{lemma}

%------------------------------------------------------------
\section{The Grassmann--Banach manifold}
\label{sec:banach:grassmann}
%------------------------------------------------------------

The Grassmannian of a Banach space is the natural parameter space for
subspaces, and it serves as the base manifold for the Tucker and
tree-based tensor manifolds.

\begin{definition}
\label{def:grassmannian}
Let $X$ be a Banach space. The \emph{Grassmann manifold} (or
\emph{Grassmannian}) of $X$ is
\begin{equation*}
    \GrassB{X} \coloneqq \{U \subset X : U \text{ is a split subspace of } X\}.
\end{equation*}
For $r \in \NNz$, the \emph{finite Grassmannian} is
\begin{equation*}
    \Grass{r}{X} \coloneqq
    \{U \in \GrassB{X} : \dim U = r\}.
\end{equation*}
\end{definition}

Both $X$ and $\{0\}$ belong to $\GrassB{X}$.
By Proposition~\ref{prop:finitedim_split}, every $r$-dimensional subspace
of $X$ belongs to $\GrassB{X}$, so $\Grass{r}{X}$ is well defined.

\begin{theorem}[Grassmann--Banach manifold, {\cite{Douady1966}}]
\label{thm:grassmann_manifold}
Let $X$ be a Banach space. The Grassmannian $\GrassB{X}$ is an analytic
Banach manifold, with atlas
$\bigl\{\GrassB{W, X},\, \Psi_{U \oplus W}\bigr\}_{U \in \GrassB{X}}$,
where for each splitting $X = U \oplus W$:
\begin{enumerate}
\item[\textup{(i)}] $\GrassB{W, X} \coloneqq
    \{V \in \GrassB{X} : X = V \oplus W\}$
    is the chart domain at $U$.
\item[\textup{(ii)}] The chart map
    $\Psi_{U \oplus W}: \GrassB{W, X} \to \cL(U, W)$
    is defined by
    \begin{equation}
    \label{eq:grassmann_chart}
        \Psi_{U \oplus W}(V) \coloneqq
        \Proj{W}{U}\big|_V \circ \bigl(\Proj{U}{W}\big|_V\bigr)^{-1},
    \end{equation}
    with inverse
    $\Psi_{U \oplus W}^{-1}(L) = G(L) \coloneqq
    \{(\Id{X} + L)(u) : u \in U\}$.
\end{enumerate}
The transition maps are analytic. Each chart domain
$\GrassB{W, X}$ is a Banach manifold modelled on $\cL(U, W)$.
\end{theorem}

The chart map $\Psi_{U \oplus W}$ sends a subspace $V \in \GrassB{W,X}$
to the linear map $L: U \to W$ whose graph $G(L)$ equals $V$. The
identity $\Psi_{U\oplus W}^{-1}(L) = \exp(L)(U)$, combined with
Proposition~\ref{prop:nilpotent_algebra}, shows that the inverse chart is
a degree-one polynomial map (hence analytic). Analyticity of the
transition maps follows by a direct computation using
Lemma~\ref{lem:common_complement}.

\begin{remark}
\label{rem:grassmann_not_modelled}
The full Grassmannian $\GrassB{X}$ is a Banach manifold but is
\emph{not} modelled on a single Banach space: if $U$ and $U'$ are
finite-dimensional with $\dim U \neq \dim U'$, then $\cL(U, W)$ and
$\cL(U', W')$ are non-isomorphic. This is not a deficiency but a
structural fact: $\GrassB{X}$ decomposes into connected components
$\GrassB{X} = \bigsqcup_{r \in \NNz} \Grass{r}{X}$, each of which
\emph{is} modelled on a single Banach space.
\end{remark}

\begin{proposition}
\label{prop:grassr_connected}
For each $r \in \NNz$, the finite Grassmannian $\Grass{r}{X}$ is a
connected component of $\GrassB{X}$, and it is an analytic Banach manifold
modelled on $\cL(U, W)$ for any $U \in \Grass{r}{X}$ and complement $W$
\textup{(}the isomorphism class of $\cL(U, W)$ does not depend on the
choice of $U$ or $W$\textup{)}.
\end{proposition}

\begin{remark}[Grassmannian of a normed space]
\label{rem:grassmann_normed}
The construction extends to normed (non-complete) spaces. Let
$(X, \|\cdot\|)$ be a normed space with completion $\bar X$. Define
$\Grass{r}{X}$ as the set of $r$-dimensional subspaces of $X$ that are
also split in $\bar X$. By Lemma~\ref{lem:normed_grassmann}, every chart
domain $\GrassB{W, \bar X}$ is contained in $\Grass{r}{X}$, so the atlas
of $\Grass{r}{\bar X}$ restricts to an analytic atlas on $\Grass{r}{X}$.
This variant is needed in Section~\ref{sec:tucker_manifold:charts}, where
the component spaces $V_\alpha$ are normed but not necessarily complete.
\end{remark}

The tangent space to $\GrassB{X}$ at $U$ is identified as follows.

\begin{proposition}
\label{prop:grassmann_tangent}
For $U \in \GrassB{X}$ with complement $W$, the tangent space
$T_U \GrassB{X} \cong \cL(U, W)$.
\end{proposition}

This identification is consistent with the finite-dimensional case: in
$\RR^n$, the tangent space to the Grassmannian $Gr(r, n)$ at a subspace
$U$ is the space of linear maps from $U$ to its orthogonal complement,
recovering the classical description.

%------------------------------------------------------------
\section{Banach manifolds}
\label{sec:banach:manifolds}
%------------------------------------------------------------

We recall the definition of a Banach manifold in its general form, since
the full Grassmannian $\GrassB{X}$ and the full tensor space
$\VDnorm$ are manifolds not modelled on a fixed Banach space.

\begin{definition}
\label{def:banach_manifold}
Let $\mathbb{M}$ be a set. An \emph{atlas of class $\mathcal{C}^p$}
(for $p \geq 0$ or $p = \infty$ or $p = \omega$, the analytic case)
on $\mathbb{M}$ is a family of \emph{charts}
$\{(M_i, u_i)\}_{i \in A}$ satisfying:
\begin{enumerate}
\item[\textup{(AT1)}] $\{M_i\}_{i\in A}$ covers $\mathbb{M}$.
\item[\textup{(AT2)}] For each $i \in A$, the chart $u_i: M_i \to U_i$
    is a bijection onto an open set $U_i$ in some Banach space $X_i$,
    and for any $i, j \in A$, the set $u_i(M_i \cap M_j)$ is open in $X_i$.
\item[\textup{(AT3)}] Whenever $M_i \cap M_j \neq \emptyset$,
    the \emph{transition map}
    $u_j \circ u_i^{-1} : u_i(M_i \cap M_j) \to u_j(M_i \cap M_j)$
    is a diffeomorphism of class $\mathcal{C}^p$.
\end{enumerate}
Two atlases are \emph{compatible} if their union is again an atlas. A
\emph{$\mathcal{C}^p$-Banach manifold} is a set $\mathbb{M}$ together
with a maximal (complete) atlas of class $\mathcal{C}^p$. If all
$X_i$ are isomorphic to a single Banach space $X$, we say $\mathbb{M}$
is \emph{modelled on $X$}.
\end{definition}

\begin{example}
\label{ex:manifold:banach_space}
Every Banach space $X$ is a $\mathcal{C}^\infty$-Banach manifold
modelled on itself, with the single chart $(X, \Id{X})$.
\end{example}

\begin{example}
\label{ex:manifold:GL}
If $X$ is a Banach space, then $\GL{X}$ is an open subset of $\cL(X)$,
and hence an analytic Banach manifold modelled on $\cL(X)$. The maps
$A \mapsto A^{-1}$ and $(A,B) \mapsto AB$ are analytic
\textup{(}see~\cite[2.7, 2.8]{Upmeier1985}\textup{)}.
\end{example}

\begin{definition}
\label{def:tangent_space}
Let $\mathbb{M}$ be a $\mathcal{C}^p$-Banach manifold with $p \geq 1$
and $m \in \mathbb{M}$. A \emph{tangent vector} at $m$ is an equivalence
class of triples $(U_i, u_i, v)$ where $(U_i, u_i)$ is a chart at $m$
and $v \in X_i$, under the equivalence
$(U_i, u_i, v) \sim (U_j, u_j, w)$ if $D(u_j \circ u_i^{-1})(u_i(m)) v = w$.
The \emph{tangent space} $T_m \mathbb{M}$ is the set of all tangent vectors
at $m$, which inherits the structure of a Banach space from any chart.
\end{definition}

\begin{definition}
\label{def:immersion}
A $\mathcal{C}^1$-map $f: \mathbb{M} \to \mathbb{N}$ between Banach
manifolds is an \emph{immersion} at $m$ if $Df(m): T_m\mathbb{M} \to
T_{f(m)}\mathbb{N}$ is injective with closed split image. It is an
\emph{immersion} if it is an immersion at every point. An
\emph{immersed submanifold} of $\mathbb{N}$ is the image of an injective
immersion.
\end{definition}

The closedness and split condition on the image of $Df(m)$ is automatic
in finite dimensions and in Hilbert spaces (where every closed subspace is
split), but is a genuine constraint in Banach spaces. This is the key
difficulty resolved in Section~\ref{sec:tucker_manifold:immersion}.

%------------------------------------------------------------
\section{Multilinear maps and Fréchet differentiability}
\label{sec:banach:multilinear}
%------------------------------------------------------------

Since the tensor product map is multilinear, the following general results
will be used repeatedly.

\begin{definition}
A map $M: X_1 \times \cdots \times X_d \to Y$ between normed spaces is
\emph{multilinear} if it is linear in each argument separately. It is
\emph{bounded} (equivalently, \emph{continuous}) if
\begin{equation*}
    \|M\| \coloneqq \sup_{\substack{x_j \in X_j \\ \|x_j\|_j \leq 1}}
    \|M(x_1, \ldots, x_d)\|_Y < \infty.
\end{equation*}
\end{definition}

The following proposition shows that continuity implies a much stronger
regularity.

\begin{proposition}[{\cite[Proposition~79]{HJ2012}}]
\label{prop:multilinear_smooth}
Let $X_1, \ldots, X_d$ and $Y$ be normed spaces and let
$M: X_1 \times \cdots \times X_d \to Y$ be a continuous multilinear map.
Then $M$ is $\mathcal{C}^\infty$-Fréchet differentiable, and
$D^k M = 0$ for all $k > d$.
\end{proposition}

The vanishing of derivatives beyond order $d$ reflects the polynomial
nature of multilinear maps. In the complex case, a stronger conclusion
holds.

\begin{proposition}[{\cite[Theorem~160]{HJ2012}}]
\label{prop:analytic_c1}
Let $X$, $Y$ be complex Banach spaces and $U \subset X$ open. A map
$f: U \to Y$ is analytic if and only if it is $\mathcal{C}^1$-Fréchet
differentiable.
\end{proposition}

\begin{corollary}
\label{cor:multilinear_analytic}
Every continuous multilinear map between complex Banach spaces is analytic.
\end{corollary}

In real Banach spaces, $\mathcal{C}^\infty$ and analytic are distinct
notions; the tensor product map is $\mathcal{C}^\infty$ but need not be
analytic over $\RR$. For the real case, $\mathcal{C}^\infty$-regularity
is sufficient for the manifold constructions in Chapters~\ref{chap:tucker_manifold}
and~\ref{chap:TB_manifold}.

\begin{remark}
\label{rem:regularity_manifold}
Proposition~\ref{prop:multilinear_smooth} is the reason why the Tucker
and tree-based tensor manifolds are $\mathcal{C}^\infty$-Banach manifolds
(and analytic over $\CC$): their chart transition maps are compositions of
continuous multilinear maps with bounded linear projections, all of which
are $\mathcal{C}^\infty$-Fréchet differentiable.
\end{remark}

\subsection*{The Grassmannian revisited: Lemma on normed spaces}

For later use in Section~\ref{sec:tucker_manifold:charts}, we record
the following lemma about Grassmannians of normed (non-complete) spaces.

\begin{lemma}
\label{lem:normed_grassmann}
Let $(X, \|\cdot\|)$ be a normed space with completion $\bar X$. Let
$U \in \Grass{r}{\bar X}$ with $U \subset X$, and suppose
$\bar X = U \oplus W$ for some $W \in \GrassB{\bar X}$. Then every
$V \in \GrassB{W, \bar X}$ satisfies $V \subset X$.
\end{lemma}

\begin{proof}
Observe that $X = U \oplus (W \cap X)$, where $W \cap X$ is dense in $W$.
Suppose for contradiction that there exists $V \in \GrassB{W, \bar X}$
with $V \not\subset X$. Since $V$ is finite-dimensional and $\bar X =
V \oplus W$, the subspace $V \cap X$ is a proper subspace of $V$. On
the other hand, $X = (V \cap X) \oplus (W \cap X)$ implies $\bar X =
(V \cap X) \oplus W$, so $V \cap X \in \GrassB{W, \bar X}$. But
$\dim(V \cap X) < \dim V = r$, contradicting the fact that every element
of $\GrassB{W, \bar X}$ has dimension $r$.
\end{proof}


%------------------------------------------------------------
\section{Curves in the Grassmannian and retractions}
\label{sec:banach:retractions}
%------------------------------------------------------------

The Tucker and tree-based tensor manifolds live over Grassmannians, and
the numerical algorithms of Appendix~\ref{app:algorithms} — the
projector-splitting integrator, the DMRG sweep, and the geometric
ALS — require the ability to move along curves in $\Grass{r}{X}$
efficiently. This section develops the necessary tools: explicit
curves in $\Grass{r}{L^2}$, the notion of retraction, and the
connection between the chart-based retraction and the exponential map.

\subsection*{Curves in the Grassmannian via the chart}

Fix $U_0 \in \Grass{r}{X}$ with complement $W_0$, $X = U_0 \oplus W_0$.
By Theorem~\ref{thm:grassmann_manifold}, a $\mathcal{C}^1$-curve
$\gamma: (-\varepsilon, \varepsilon) \to \Grass{r}{X}$ with
$\gamma(0) = U_0$ lying in the chart domain $\GrassB{W_0, X}$ is
uniquely represented by a $\mathcal{C}^1$-curve
$L: (-\varepsilon, \varepsilon) \to \cL(U_0, W_0)$ with $L(0) = 0$
and $\gamma(t) = G(L(t)) = (\Id{X} + L(t))(U_0)$.
The velocity at $t=0$ is $\dot\gamma(0) = \dot L(0) \in \cL(U_0, W_0)
= T_{U_0}\Grass{r}{X}$.

\begin{example}[Curves in $\Grass{1}{L^2({[0,1]})}$]
\label{ex:grassmann_curves}
Let $X = L^2([0,1])$, $r=1$, $U_0 = \spann\{\phi_1\}$, where
$\phi_k(x) = \sqrt{2}\sin(k\pi x)$ are the Dirichlet eigenfunctions of
$-d^2/dx^2$.

\paragraph{Rotation between eigenfunctions.}
For fixed $k \geq 2$, the curve
\begin{equation*}
    \gamma(t) = \spann\{\cos(t)\phi_1 + \sin(t)\phi_k\},
    \quad t \in (-\pi/2, \pi/2),
\end{equation*}
is $\mathcal{C}^\infty$ in $\Grass{1}{L^2}$. Its chart coordinate is
$L(t) = \tan(t)\phi_k$ (in $\cL(\spann\{\phi_1\}, W_0)$, identifying
$\phi_k$ with the map $\lambda\phi_1 \mapsto \lambda\phi_k$), and
$\dot L(0) = \phi_k$.

\paragraph{Propagating Gaussian support.}
Define $\psi_a(x) = C_a e^{-a(x-1/2)^2}$ (normalised). The curve
$t \mapsto \spann\{\psi_{a+t}\}$ traces the evolution of the support
subspace as the Gaussian width changes. At $t=0$, the velocity
$\dot L(0) = \frac{d}{dt}\big|_0 \Psi_{U_0\oplus W_0}(\spann\{\psi_{a+t}\})
\in \cL(\spann\{\psi_a\}, \{\psi_a\}^\perp)$ encodes the instantaneous
rate of change of the dominant mode.
\end{example}

\begin{example}[Curves in $\Grass{r}{L^2}$ for $r=2$]
\label{ex:grassmann_curves_r}
For $r=2$ and $U_0 = \spann\{\phi_1,\phi_2\}$, the curve
\begin{equation*}
    \gamma(t) = \spann\bigl\{
    \cos(t)\phi_1 + \sin(t)\phi_3,\;
    \cos(t)\phi_2 + \sin(t)\phi_4
    \bigr\}
\end{equation*}
has chart matrix $[L(t)]_{ij} = \tan(t)\delta_{ij}$: a uniform
rotation of both basis vectors. For the Tucker manifold
$\Tucker{(2,2)}{L^2 \otimes L^2}$, a pair of such curves (one per
factor) gives a curve in the base $\Grass{2}{L^2} \times \Grass{2}{L^2}$
of the Tucker bundle.
\end{example}

\subsection*{Retractions on Banach manifolds}

\begin{definition}
\label{def:retraction}
A \emph{retraction} on a $\mathcal{C}^2$-Banach manifold $\mathbb{M}$
is a $\mathcal{C}^1$-map $R: T\mathbb{M} \to \mathbb{M}$ such that,
for every $m \in \mathbb{M}$, the restriction $R_m \coloneqq
R|_{T_m\mathbb{M}}$ satisfies:
\begin{enumerate}
\item[\textup{(R1)}] $R_m(0_m) = m$.
\item[\textup{(R2)}] $DR_m(0_m) = \Id{T_m\mathbb{M}}$.
\end{enumerate}
\end{definition}

Condition~(R2) says the retraction is a first-order approximation to
the exponential map. Retractions are the key ingredient for geometric
optimisation (gradient descent, Newton's method) and geometric
integration on manifolds; they replace the exponential map whenever
no Riemannian metric is available.

\begin{example}[Retraction on $\Grass{r}{X}$]
\label{ex:grassmann_retraction}
For $U \in \Grass{r}{X}$ and $L \in T_U\Grass{r}{X} = \cL(U, W)$:
\begin{equation}
\label{eq:grassmann_retraction}
    R_U(L) \coloneqq (\Id{X} + L)(U).
\end{equation}
Conditions~(R1)--(R2) are immediate from the chart map
(Theorem~\ref{thm:grassmann_manifold}).
\end{example}

\begin{remark}[Retraction vs.\ geodesic]
\label{rem:retraction_vs_exp}
In the Hilbert case, the geodesic on $\Grass{r}{H}$ from $U$ with
velocity $L \in \cL(U, U^\perp)$ is
$\gamma(t) = \cos(t|L|)(U) \oplus \sin(t|L|)\mathrm{Im}(L/|L|)$,
where $|L| = (L^*L)^{1/2}$. The chart-based retraction~\eqref{eq:grassmann_retraction}
equals the geodesic to first order in $t$ (agreeing at $t=0$ and having
the same velocity). For $r=1$, the geodesic is the great circle on
$\mathbb{P}(H)$ — the Rabi oscillation in quantum mechanics.

In the non-Hilbert Banach setting, there is no canonical Riemannian
metric; the chart-based retraction is the natural substitute and is
used in the projector-splitting
integrator~\cite{LubichOseledetsVandereycken2015}.
\end{remark}

\begin{example}[Retraction on the Tucker manifold]
\label{ex:tucker_retraction}
For $\bv \in \Tucker{\fr}{\VD}$ and $\dot\bv \in \ZD{\bv}$, writing
$\dot\bv = d\Phi_\bv^{-1}[\dot\bv]$ in chart coordinates
$((\delta L_\alpha)_\alpha, \delta C^{(D)})$:
\begin{equation}
\label{eq:tucker_retraction}
    R_\bv(\dot\bv)
    = \sum_{(i_\alpha)}
    (C^{(D)} + \delta C^{(D)})_{(i_\alpha)}
    \bigotimes_\alpha
    (\Id{} + \delta L_\alpha)(u^{(\alpha)}_{i_\alpha}).
\end{equation}
This satisfies (R1)--(R2) by construction and is the retraction used
in the geometric ALS and DMRG algorithms of
Appendix~\ref{app:algorithms}.
\end{example}

%------------------------------------------------------------
\section{Historical notes}
\label{sec:banach:history}
%------------------------------------------------------------

The material of this chapter spans several decades of functional
analysis and infinite-dimensional differential geometry. We trace the
origins of the main topics: Banach space theory, complemented subspaces,
the Grassmann--Banach manifold, the general theory of Banach manifolds,
and the notion of retraction.

\paragraph{Banach spaces and the geometry of infinite-dimensional spaces.}
The systematic study of complete normed spaces was inaugurated by Stefan
Banach in his 1922 doctoral thesis and codified in the monograph
\emph{Théorie des opérations linéaires}~\cite{Banach1932}. The three
fundamental theorems of linear functional analysis — the Hahn--Banach
theorem, the open mapping theorem, and the uniform boundedness principle
— were all established or placed on definitive footing in this period.
The question of when a closed subspace is complemented (i.e., has a
bounded projection from the ambient space) was raised immediately:
Banach observed that every finite-dimensional or finite-codimensional
closed subspace of a Banach space is complemented, and the first deep
negative result — that $c_0$ is not complemented in $\ell^\infty$ —
is due to Phillips~\cite{Phillips1940}. The landmark result that $\ell^1$
is the only separable Banach space in which every closed subspace is
complemented is due to Lindenstrauss and
Tzafriri~\cite{LindenstraussTzafriri1971}, answering a long-standing
question of Banach. The modern geometric theory of Banach spaces,
encompassing type and cotype, uniform convexity, and local theory, was
developed primarily by Mazur, Pe\l czy\'{n}ski, Lindenstrauss, and
Milman from the 1950s onwards; the monograph of Lindenstrauss and
Tzafriri~\cite{LindenstraussTzafriri1977} remains the standard reference.

\paragraph{Complemented subspaces and projections.}
The notion of complemented subspace
(Definition~\ref{def:complemented_subspace}) and the basic properties of
bounded projections (Section~\ref{sec:banach:complemented}) are classical;
see Lindenstrauss--Tzafriri~\cite{LindenstraussTzafriri1977} or the
monograph of Megginson~\cite{Megginson1998} for a modern treatment.
The specific form of the projection $\Proj{U}{W}$ associated to a direct
sum decomposition $X = U \oplus W$
(Proposition~\ref{prop:projection_characterisation}) and the
norm-continuity of the map $L \mapsto (\Id{} + L)(U)$ used in the
Grassmannian charts are standard results; the presentation here follows
Douady~\cite{Douady1966}.

\paragraph{The Grassmann--Banach manifold.}
The construction of the Grassmannian of a Banach space as an
infinite-dimensional analytic manifold
(Theorem~\ref{thm:grassmann_manifold}) is due to Douady~\cite{Douady1966},
who established the analytic atlas using the graph construction and
proved analyticity of the transition maps. An independent and closely
related construction appears in the work of Porta and
Recht~\cite{PortaRecht1994}, who studied the geometry of the Grassmannian
in the context of operator algebras, and of Corach, Porta, and
Recht~\cite{CorachPortaRecht1993}, who developed the metric geometry and
geodesics of the Grassmannian of a $C^*$-algebra. These works established
the Grassmannian as a symmetric space in the Hilbert-module setting and
identified the geodesics as arcs of the unitary group, giving the formula
recalled in Remark~\ref{rem:retraction_vs_exp}. The finite-dimensional
case — the classical Grassmannian $Gr(r, n)$ as a homogeneous space of
$GL(n)$ — is treated in every textbook on differential geometry;
see Milnor--Stasheff~\cite{MilnorStasheff1974} for the characteristic
class theory.

\paragraph{Banach manifolds.}
The foundations of the theory of manifolds modelled on Banach spaces were
laid independently by several authors in the late 1950s and early 1960s.
The definitive early reference is the monograph of
Lang~\cite{Lang1999}, which establishes the basic theory of Banach
manifolds — charts, atlases, tangent bundles, vector fields, and the
Frobenius theorem — in the generality required for infinite-dimensional
applications. A parallel development, closer to the needs of global
analysis, appears in the work of Eells~\cite{Eells1966} and in the
monograph of Abraham, Marsden, and Ratiu~\cite{AbrahamMarsdenRatiu1988},
which provides the most thorough treatment of infinite-dimensional
manifolds available and covers connections, Riemannian geometry, and
Hamiltonian mechanics in the Banach setting. The foundational text of
Dieudonné~\cite{Dieudonne1972} treats the local differential calculus
on Banach spaces in full generality, establishing the chain rule,
the inverse function theorem, and the implicit function theorem
(Theorem~\ref{thm:IFT}) in the form used in Chapters~\ref{chap:tucker_manifold}
and~\ref{chap:TB_manifold}.

\paragraph{Fréchet differentiability and multilinear maps.}
The Fréchet derivative in Banach spaces
(Section~\ref{sec:banach:multilinear}) is the canonical notion of
differentiability for maps between infinite-dimensional spaces; it agrees
with the Gâteaux derivative for maps whose Gâteaux derivative is bounded
and continuous, which is always the case in our applications. The
identification of bounded multilinear maps with bounded linear maps on
tensor products (the universal property, Proposition~\ref{prop:multilinear_universal})
is classical; see Dieudonné~\cite{Dieudonne1972}, Chapter~12, or
Ryan~\cite{Ryan2002} for a treatment aimed at Banach space theory.
The $\mathcal{C}^\infty$ regularity of the tensor product map
(Proposition~\ref{prop:tensormap_smooth}), which is used throughout
Chapters~\ref{chap:tucker_manifold}--\ref{chap:TB_manifold}, follows
immediately from its polynomial (hence real-analytic) nature.

\paragraph{Retractions and geometric optimisation.}
The notion of retraction on a Riemannian manifold as a first-order
approximation to the exponential map
(Definition~\ref{def:retraction}) was formalised for optimisation
purposes by Shub~\cite{Shub1986} and systematised by Absil, Mahony, and
Sepulchre~\cite{AbsilMahonySepulchre2008} in their monograph on
optimisation on matrix manifolds. The specific chart-based retraction on
the Grassmannian (Example~\ref{ex:grassmann_retraction}) is the one
used in practice in geometric integration algorithms; the
projector-splitting integrator of Lubich, Oseledets, and
Vandereycken~\cite{LubichOseledetsVandereycken2015} employs exactly
this retraction at each level of the tree. In the Hilbert case the
relation between the chart retraction and the geodesic (great circle) is
well known; the Banach case requires a different approach since no
canonical Riemannian metric is available, and the chart retraction is
the natural substitute.
