%============================================================
%  ch10_DF.tex
%  Chapter 10: The Dirac--Frenkel Variational Principle
%
%  Tensor Formats in Banach Spaces
%  Antonio Falcó, CEU Universities
%
%  Based on: FHN2019 (Sec. 5), FHN2023 (Sec. 5)
%  Estimated length: ~35 pp.
%============================================================

\chapter{The Dirac--Frenkel Variational Principle}
\label{chap:DF}

The geometric framework developed in Chapters~\ref{chap:tucker_manifold}
and~\ref{chap:TB_manifold} has a direct dynamical application: it
provides the rigorous foundation for \emph{model reduction} of
differential equations on tensor Banach spaces. Given an abstract
Cauchy problem on $\VDnorm$, one looks for an approximate solution
that evolves on the manifold $\Tucker{\fr}{\VD}$ (or $\FTv$).
The natural way to constrain the dynamics to the manifold is the
\emph{Dirac--Frenkel variational principle}: at each time $t$, the
velocity $\dot{\bv}_r(t)$ is chosen as the best approximation of the
vector field $\bF(t, \bv_r(t))$ in the tangent space $\ZD{\bv_r(t)}$.

This principle was originally proposed by Dirac and Frenkel in 1930 for
the time-dependent Schrödinger equation in Hilbert spaces. The present
chapter extends it to general Banach tensor spaces, establishing the
existence of the reduced ODE and characterising its solutions.
Sections~\ref{sec:DF:cauchy}--\ref{sec:DF:hilbert} set up the
abstract framework. Section~\ref{sec:DF:tucker} proves the Tucker
version (Theorems~5.2 and~5.4 of~\cite{FHN2019}).
Section~\ref{sec:DF:treebased} extends to the tree-based setting
(\cite{FHN2023}).

%------------------------------------------------------------
\section{The abstract Cauchy problem on tensor Banach spaces}
\label{sec:DF:cauchy}
%------------------------------------------------------------

Let $(V_\alpha, \|\cdot\|_\alpha)$ be normed spaces for $\alpha \in D$,
and let $\VDnorm = {}_{\|\cdot\|_D}\bigotimes_{\alpha \in D} V_\alpha$
be a reflexive Banach tensor space satisfying condition~\condINJ.
Consider the abstract Cauchy problem:
\begin{align}
    \dot{\bu}(t) &= \bF(t, \bu(t)), \quad t \geq 0,
    \label{eq:DF:ode} \\
    \bu(0) &= \bu_0,
    \label{eq:DF:ic}
\end{align}
where $\bu_0 \neq \mathbf{0}$, and $\bF: [0, \infty) \times \VDnorm
\to \VDnorm$ satisfies the usual conditions for existence and uniqueness
of solutions (e.g., Lipschitz continuity in the second argument).

The goal is to approximate $\bu(t)$ by a \emph{differentiable curve}
$t \mapsto \bv_r(t)$ lying in $\Tucker{\fr}{\VD}$ for some fixed
$\fr \in \AD{\VD}$ with $\fr \neq \mathbf{0}$, with initial condition
$\bv_r(0) = \bv_0 \in \Tucker{\fr}{\VD}$ chosen as an approximation
of $\bu_0$.

\begin{remark}
\label{rem:DF:initial}
By Theorem~\ref{thm:best_approx_tucker}, a best approximation
$\bv_0 \in \Tucker{\leq\fr}{\VD}$ of $\bu_0$ exists when $\VDnorm$
is reflexive. The initial condition $\bv_0 \in \Tucker{\fr}{\VD}$
(with exact rank $\fr$) can always be arranged by taking
$\bv_0 \in \Tucker{\fr'}{\VD}$ for some $\fr' \leq \fr$ and
adjusting the rank.
\end{remark}

%------------------------------------------------------------
\section{The principle in Hilbert spaces}
\label{sec:DF:hilbert}
%------------------------------------------------------------

When $\VDnorm$ is a Hilbert space with inner product
$\langle \cdot, \cdot \rangle_D$, the constraint to $\Tucker{\fr}{\VD}$
takes the classical form. Since $\ZD{\bv_r(t)} \in \GrassB{\VDnorm}$
(Theorem~\ref{thm:tucker_immersion}), the orthogonal projection
$\Pmetric{\bv_r(t)}$ onto $\ZD{\bv_r(t)}$ is well-defined and the
\emph{Dirac--Frenkel equation} reads:
\begin{equation}
\label{eq:DF:hilbert}
    \dot{\bv}_r(t) = \Pmetric{\bv_r(t)}\bigl(\bF(t, \bv_r(t))\bigr),
\end{equation}
which is equivalent to the variational condition:
\begin{equation}
\label{eq:DF:variational}
    \langle \dot{\bv}_r(t) - \bF(t, \bv_r(t)),\, \dot{\bv} \rangle_D = 0
    \quad \text{for all } \dot{\bv} \in \ZD{\bv_r(t)}.
\end{equation}
The MCTDH (Multi-Configuration Time-Dependent Hartree) method of
quantum dynamics~\cite{Lubich2008} is precisely~\eqref{eq:DF:hilbert}
with $\VDnorm = L^2(\RR^{3d})$ and $\Tucker{\fr}{\VD}$ the Hartree
manifold ($\fr = (1,\ldots,1)$).

%------------------------------------------------------------
\section{Projections in Banach spaces}
\label{sec:DF:projection}
%------------------------------------------------------------

In a general Banach space, there is no canonical inner product and hence
no canonical projection. Two notions replace the orthogonal projection.

\subsection*{Metric projection}

\begin{definition}
\label{def:metric_projection}
For $\bv_r(t) \in \Tucker{\fr}{\VD}$ and $t \in I$, the
\emph{metric projection} $\Pmetric{\bv_r(t)}$ maps
$\dot{\bu} \in \VDnorm$ to the nearest point in $\ZD{\bv_r(t)}$:
\begin{equation*}
    \Pmetric{\bv_r(t)}(\dot{\bu})
    \coloneqq \argmin_{\dot{\bv} \in \ZD{\bv_r(t)}}
    \|\dot{\bu} - \dot{\bv}\|_D.
\end{equation*}
\end{definition}

By Corollary~\ref{cor:best_proj_tangent}, when $\VDnorm$ is reflexive
the metric projection exists. When $\VDnorm$ is strictly convex it is
unique (since $\ZD{\bv_r(t)}$ is convex). The metric projection is,
in general, a \emph{nonlinear} operator on Banach spaces.

\subsection*{Normalised duality mapping}

The duality mapping $J: \VDnorm \to 2^{\dual{\VDnorm}}$ is defined by
\begin{equation*}
    J(\bu) \coloneqq \bigl\{f \in \dual{\VDnorm} :
    \langle \bu, f \rangle = \|\bu\|_D^2 = \|f\|_*^2\bigr\}.
\end{equation*}
It is single-valued when $\VDnorm$ is smooth (Definition~\ref{def:smooth}),
and uniformly continuous on bounded sets when $\VDnorm$ is uniformly smooth.
In a Hilbert space, $J$ coincides with the Riesz isomorphism.

\begin{theorem}[Metric projection characterisation,
{\cite[Theorem~5.2]{FHN2019}}]
\label{thm:metric_projection}
Let $\VDnorm$ be reflexive and strictly convex, and assume \condINJ.
For $\bv_r(t) \in \Tucker{\fr}{\VD}$ and fixed $t \in I$, the
following are equivalent:
\begin{enumerate}
\item[\textup{(a)}] $\dot{\bv}_r(t) = \Pmetric{\bv_r(t)}
    (\bF(t,\bv_r(t)))$.
\item[\textup{(b)}] $\langle \dot{\bv}_r(t) - \dot{\bv},\,
    J(\bF(t,\bv_r(t)) - \dot{\bv}_r(t)) \rangle \geq 0$
    for all $\dot{\bv} \in \ZD{\bv_r(t)}$.
\item[\textup{(c)}] $\langle \dot{\bv},\,
    J(\bF(t,\bv_r(t)) - \dot{\bv}_r(t)) \rangle = 0$
    for all $\dot{\bv} \in \ZD{\bv_r(t)}$.
\end{enumerate}
\end{theorem}

\begin{proof}
This is a direct consequence of the classical characterisation of the
metric projection onto a closed convex set in a reflexive, strictly
convex Banach space, combined with~\cite[Proposition~2.10]{Alber1996}.
Condition~(c) recovers the Hilbert-space variational form: when $J$
is the Riesz map, (c) becomes the orthogonality condition~\eqref{eq:DF:variational}.
\end{proof}

\subsection*{Generalised projection}

When $\VDnorm$ is additionally smooth, an alternative projection is
available. Define $\phi: \VDnorm \times \VDnorm \to \RR$ by
\begin{equation*}
    \phi(\bu, \bv) = \|\bu\|_D^2 - 2\langle \bu, J(\bv) \rangle + \|\bv\|_D^2.
\end{equation*}
The \emph{generalised projection} $\Pi_{\bv_r(t)}$ minimises $\phi$
over $\ZD{\bv_r(t)}$:
\begin{equation*}
    \Pi_{\bv_r(t)}(\dot{\bu})
    \coloneqq \argmin_{\dot{\bv} \in \ZD{\bv_r(t)}}
    \phi(\dot{\bv}, \dot{\bu}).
\end{equation*}
It coincides with the metric projection when $\VDnorm$ is a Hilbert
space.

%------------------------------------------------------------
\section{The Dirac--Frenkel principle for the Tucker manifold}
\label{sec:DF:tucker}
%------------------------------------------------------------

\begin{theorem}[Dirac--Frenkel for Tucker, {\cite[Theorems~5.2,~5.4]{FHN2019}}]
\label{thm:DF_tucker}
Let $\VDnorm$ be reflexive and strictly convex, and assume \condINJ.
Let $\fr \in \AD{\VD}$ with $\fr \neq \mathbf{0}$, and let
$\bv_0 \in \Tucker{\fr}{\VD}$.
\begin{enumerate}
\item[\textup{(Existence)}] There exists a differentiable curve
    $t \mapsto \bv_r(t)$ in $\Tucker{\fr}{\VD}$, defined for $t$ in a
    neighbourhood of $0$, satisfying
    \begin{equation}
    \label{eq:DF_tucker}
        \dot{\bv}_r(t) = \Pmetric{\bv_r(t)}\bigl(\bF(t, \bv_r(t))\bigr),
        \qquad \bv_r(0) = \bv_0.
    \end{equation}

\item[\textup{(Characterisation)}] The curve $t \mapsto \bv_r(t)$
    satisfies~\eqref{eq:DF_tucker} if and only if $\dot{\bv}_r(t)
    \in \ZD{\bv_r(t)}$ and
    \begin{equation}
    \label{eq:DF_tucker_char}
        \|\dot{\bv}_r(t) - \bF(t, \bv_r(t))\|_D
        = \min_{\dot{\bv} \in \ZD{\bv_r(t)}}
        \|\dot{\bv} - \bF(t, \bv_r(t))\|_D.
    \end{equation}
\end{enumerate}
\end{theorem}

\begin{proof}[Sketch]
\textbf{Existence.} The reduced ODE~\eqref{eq:DF_tucker} is an ODE on
the Banach manifold $\Tucker{\fr}{\VD}$. By
Theorem~\ref{thm:tucker_immersion}, $\ZD{\bv_r(t)}$ is a split
subspace of $\VDnorm$ for each $\bv_r(t) \in \Tucker{\fr}{\VD}$.
The right-hand side of~\eqref{eq:DF_tucker} depends smoothly on
$\bv_r(t)$ (the metric projection onto a smoothly-varying split
subspace is smooth), so local existence and uniqueness follow from the
ODE existence theorem on Banach manifolds~\cite[Theorem~4.1.5]{Lang1999}.

\textbf{Characterisation.} Since $\ZD{\bv_r(t)}$ is closed and
convex, Corollary~\ref{cor:best_proj_tangent} gives existence of the
minimum in~\eqref{eq:DF_tucker_char}. The equivalence with the metric
projection condition~\eqref{eq:DF_tucker} follows from the uniqueness
(strict convexity) and Theorem~\ref{thm:metric_projection}.
\end{proof}

%------------------------------------------------------------
\section{Kinematic equations for the Tucker manifold}
\label{sec:DF:kinematic}
%------------------------------------------------------------

The Dirac--Frenkel ODE~\eqref{eq:DF_tucker} is an abstract equation
on the Banach manifold $\Tucker{\fr}{\VD}$. In this section we derive
its explicit form in local chart coordinates, obtaining a coupled
system of ODEs for the factor matrices and the core tensor. This
system is the Banach-space analogue of the MCTDH equations of
quantum dynamics.

\subsection*{Setup and notation}

Fix $\bv_r(t) \in \Tucker{\fr}{\VD}$ and choose at each time $t$:
\begin{itemize}
\item Bases $\{u^{(\alpha)}_{i_\alpha}(t)\}_{i_\alpha=1}^{r_\alpha}$
    of $\Umin{\alpha}{\bv_r(t)}$ for each $\alpha \in D$, normalised
    so that the Gram matrices
    $G^{(\alpha)}_{ij}(t) \coloneqq
    \langle u^{(\alpha)}_i(t), u^{(\alpha)}_j(t)\rangle_\alpha$
    satisfy $G^{(\alpha)}(t) = I_{r_\alpha}$ when $V_\alpha$ is a
    Hilbert space (orthonormal basis), or more generally
    $G^{(\alpha)}(t) = I_{r_\alpha}$ modulo a choice of
    normalisation adapted to the Banach space structure.
\item The core tensor $C^{(D)}(t) \in \RR^{\times r_\alpha}_*$
    uniquely determined by
    $\bv_r(t) = \sum_{(i_\alpha)} C^{(D)}_{(i_\alpha)}(t)
    \bigotimes_\alpha u^{(\alpha)}_{i_\alpha}(t)$.
\end{itemize}
We also define, for each $\alpha \in D$ and $i_\alpha = 1,\ldots,r_\alpha$:
\begin{equation}
\label{eq:DF:Ualpha}
    \bU^{(\alpha)}_{i_\alpha}(t)
    \coloneqq \sum_{(i_\beta)_{\beta \neq \alpha}}
    C^{(D)}_{(i_\alpha, (i_\beta)_{\beta \neq \alpha})}(t)
    \bigotimes_{\beta \neq \alpha} u^{(\beta)}_{i_\beta}(t)
    \in \Umin{D \setminus \{\alpha\}}{\bv_r(t)},
\end{equation}
and the \emph{mean-field matrix}
\begin{equation}
\label{eq:DF:meanfield}
    F^{(\alpha)}_{i_\alpha j_\alpha}(t)
    \coloneqq
    \bigl\langle
        \bU^{(\alpha)}_{i_\alpha}(t),\;
        \bigl(\Id{\alpha} \otimes
        T_\psi^{[\alpha]}\bigr)(\bF(t, \bv_r(t)))
        [\psi = u^{(\alpha)}_{j_\alpha}(t)]
    \bigr\rangle_{[\alpha]},
\end{equation}
where $T_\psi^{[\alpha]}(\bw) \coloneqq
(\psi \otimes \Id{[\alpha]})(\bw) \in \Vbigalpha^{[\alpha]}$
denotes the contraction of $\bw$ with $\psi \in V_\alpha^*$
in the $\alpha$-direction.

\subsection*{Derivation of the kinematic system}

By the tangent space identification
(Proposition~\ref{prop:tangent_space}), any velocity
$\dot{\bv}_r(t) \in \ZD{\bv_r(t)}$ can be written uniquely as
\begin{equation}
\label{eq:DF:tangent_decomp}
    \dot{\bv}_r(t)
    = \sum_{(i_\alpha)} \dot{C}^{(D)}_{(i_\alpha)}(t)
      \bigotimes_\alpha u^{(\alpha)}_{i_\alpha}(t)
    + \sum_{\alpha \in D} \sum_{i_\alpha=1}^{r_\alpha}
      \dot{u}^{(\alpha)}_{i_\alpha}(t) \otimes \bU^{(\alpha)}_{i_\alpha}(t),
\end{equation}
where $\dot{u}^{(\alpha)}_{i_\alpha}(t) \in \Wmin{\alpha}{\bv_r(t)}$
(the complement of $\Umin{\alpha}{\bv_r(t)}$ in $V_\alpha$).

The Dirac--Frenkel condition
(Theorem~\ref{thm:metric_projection}(a)$\Leftrightarrow$(c)):
\begin{equation}
\label{eq:DF:variational_banach}
    \bigl\langle \xi,\;
    J\bigl(\bF(t,\bv_r(t)) - \dot{\bv}_r(t)\bigr)
    \bigr\rangle = 0
    \quad \text{for all } \xi \in \ZD{\bv_r(t)},
\end{equation}
must hold for all basis elements of $\ZD{\bv_r(t)}$. Testing
against each type of basis element yields two families of equations.

\medskip
\noindent\textbf{Equations for the core tensor.}

Testing~\eqref{eq:DF:variational_banach} against
$\xi = \bigotimes_\alpha u^{(\alpha)}_{j_\alpha}(t)$ (the core basis
elements) gives, for each multi-index $(j_\alpha)_{\alpha \in D}$:
\begin{equation}
\label{eq:DF:core_eq}
    \dot{C}^{(D)}_{(j_\alpha)}(t)
    = \bigl\langle
        \bigotimes_\alpha u^{(\alpha)}_{j_\alpha}(t),\;
        J\bigl(\bF(t,\bv_r(t))\bigr)
      \bigr\rangle
    - \sum_\alpha \sum_{i_\alpha}
      \bigl\langle
        \bU^{(\alpha)}_{i_\alpha}(t) \otimes
        u^{(\alpha)}_{j_\alpha}(t),\;
        J(\dot{u}^{(\alpha)}_{i_\alpha}(t))
      \bigr\rangle
      C^{(D)}_{\ldots i_\alpha \ldots}(t),
\end{equation}
where $J: \VDnorm \to (\VDnorm)^*$ is the duality mapping. In a
Hilbert space (where $J$ is the Riesz isomorphism and
$\langle \cdot, J(\cdot)\rangle = \langle \cdot, \cdot\rangle$),
this reduces to
\begin{equation}
\label{eq:DF:core_hilbert}
    \dot{C}^{(D)}_{(j_\alpha)}(t)
    = \bigl\langle
        \bF(t,\bv_r(t)),\;
        \bigotimes_\alpha u^{(\alpha)}_{j_\alpha}(t)
      \bigr\rangle_D
    - \sum_\alpha \sum_{i_\alpha}
      \langle \dot{u}^{(\alpha)}_{i_\alpha}, u^{(\alpha)}_{j_\alpha}\rangle_\alpha
      \, C^{(D)}_{\ldots i_\alpha \ldots}(t).
\end{equation}

\medskip
\noindent\textbf{Equations for the basis vectors.}

Testing~\eqref{eq:DF:variational_banach} against
$\xi = w \otimes \bU^{(\alpha)}_{j_\alpha}(t)$ for each
$\alpha \in D$, $j_\alpha \in \{1,\ldots,r_\alpha\}$, and
$w \in \Wmin{\alpha}{\bv_r(t)}$ gives:
\begin{equation}
\label{eq:DF:basis_eq}
    \bigl\langle
        w \otimes \bU^{(\alpha)}_{j_\alpha}(t),\;
        J\bigl(\bF(t,\bv_r(t)) - \dot{\bv}_r(t)\bigr)
    \bigr\rangle = 0
    \quad \forall\, w \in \Wmin{\alpha}{\bv_r(t)}.
\end{equation}
This can be rewritten, using the tensor product structure and the
mean-field matrix~\eqref{eq:DF:meanfield}, as:
\begin{equation}
\label{eq:DF:basis_explicit}
    P^{(\alpha)}_{W}(t)\,
    \Bigl[
        \sum_{j_\alpha} F^{(\alpha)}_{j_\alpha \cdot}(t)\,
        u^{(\alpha)}_{j_\alpha}(t)
        - \sum_{j_\alpha} M^{(\alpha)}_{j_\alpha \cdot}(t)\,
        \dot{u}^{(\alpha)}_{j_\alpha}(t)
    \Bigr] = 0,
\end{equation}
where $P^{(\alpha)}_W(t) = \Proj{\Wmin{\alpha}{\bv_r(t)}}{\Umin{\alpha}{\bv_r(t)}}$
is the projection onto the complement, and
$M^{(\alpha)}_{ij}(t) \coloneqq
\langle \bU^{(\alpha)}_i(t), \bU^{(\alpha)}_j(t)\rangle_{[\alpha]}$
is the \emph{mass matrix} (the Gram matrix of the $\bU^{(\alpha)}_{i_\alpha}$ vectors).

\begin{remark}[Gauge freedom and the MCTDH constraint]
\label{rem:gauge}
The kinematic equations~\eqref{eq:DF:core_eq}
and~\eqref{eq:DF:basis_explicit} are underdetermined: one can add
to $\dot{u}^{(\alpha)}_{i_\alpha}$ any vector in $\Umin{\alpha}{\bv_r(t)}$
without changing $\dot{\bv}_r(t)$ (since such a change stays within
the core subspace and is absorbed by $\dot{C}^{(D)}$). This is the
\emph{gauge freedom} of the Tucker format
(Remark~\ref{rem:gauge_freedom} in Chapter~\ref{chap:algTBT}).

The standard choice to remove this degeneracy is the
\emph{tangential constraint}:
\begin{equation}
\label{eq:DF:tangential}
    \langle \dot{u}^{(\alpha)}_{i_\alpha}(t),
    u^{(\alpha)}_{j_\alpha}(t) \rangle_\alpha = 0
    \quad \text{for all } i_\alpha, j_\alpha \in \{1,\ldots,r_\alpha\},
    \; \alpha \in D,
\end{equation}
which forces $\dot{u}^{(\alpha)}_{i_\alpha}(t) \in
\Wmin{\alpha}{\bv_r(t)}$ (orthogonal to the current subspace).
In the Hilbert case, this is the MCTDH tangential constraint
of~\cite{KochLubich2007}.
\end{remark}

\subsection*{The kinematic system in the Hilbert case}

In a Hilbert tensor space $\VDnorm = \bigotimes_j H_j$ with
orthonormal bases $\{u^{(\alpha)}_{i_\alpha}(t)\}$, imposing the
tangential constraint~\eqref{eq:DF:tangential} gives the closed system:

\begin{subequations}
\label{eq:DF:MCTDH}
\begin{align}
    \dot{C}^{(D)}_{(j_\alpha)}(t)
    &= \bigl\langle \bF(t,\bv_r(t)),
       \bigotimes_\alpha u^{(\alpha)}_{j_\alpha}(t)\bigr\rangle_D,
    \label{eq:DF:MCTDH_core}\\[4pt]
    \dot{u}^{(\alpha)}_{j_\alpha}(t)
    &= P^{(\alpha)}_W(t)
       \bigl[M^{(\alpha)}(t)^{-1}\bigr]_{j_\alpha \cdot}
       \sum_{i_\alpha} F^{(\alpha)}_{i_\alpha \cdot}(t)\,
       u^{(\alpha)}_{i_\alpha}(t),
    \label{eq:DF:MCTDH_basis}
\end{align}
\end{subequations}
where
\begin{equation}
\label{eq:DF:mean_field_hilbert}
    F^{(\alpha)}_{i_\alpha j_\alpha}(t)
    = \bigl\langle \bU^{(\alpha)}_{i_\alpha}(t),
        (\Id{\alpha} \otimes
        \langle u^{(\alpha)}_{j_\alpha}(t), \cdot\rangle_\alpha)
        \bF(t,\bv_r(t))
    \bigr\rangle_{[\alpha]}.
\end{equation}

\begin{remark}[Connection to MCTDH]
\label{rem:MCTDH_connection}
Equations~\eqref{eq:DF:MCTDH} are precisely the
\emph{Multi-Configuration Time-Dependent Hartree} (MCTDH) equations
of~\cite{KochLubich2007}, generalised from the rank-one case
$\fr = \mathbf{1}$ to arbitrary Tucker rank $\fr$.
For $\fr = \mathbf{1}$ (Hartree manifold, Example~\ref{ex:hartree}):
\begin{itemize}
\item The core reduces to a scalar $C^{(D)}(t) = \lambda(t) \in \RR^*$,
    and~\eqref{eq:DF:MCTDH_core} gives
    $\dot\lambda = \langle \bF, \bigotimes_\alpha v_\alpha\rangle_D$.
\item The mean-field matrix $F^{(\alpha)}$ reduces to a scalar
    $F^{(\alpha)} = \langle \bu_0, (\Id\otimes\langle v_\alpha,\cdot\rangle)\bF\rangle$
    (the standard mean-field force), and~\eqref{eq:DF:MCTDH_basis}
    gives $\dot{v}_\alpha = \lambda^{-1} P_W^{(\alpha)} F^{(\alpha)}$,
    recovering the classical Hartree equations.
\end{itemize}
\end{remark}

\begin{remark}[Banach case: the duality mapping replaces the inner product]
\label{rem:kinematic_banach}
In a non-Hilbert Banach tensor space, the kinematic equations
carry through with the duality mapping $J$ in place of the inner product.
The core equation~\eqref{eq:DF:core_eq} involves $\langle \cdot, J(\cdot)\rangle$
instead of $\langle\cdot,\cdot\rangle_D$. The basis equations involve the
duality pairing between $V_\alpha$ and $V_\alpha^*$, which replaces the
inner product. The mass matrix $M^{(\alpha)}$ must be replaced by the
\emph{duality mass matrix}
$M^{(\alpha)}_{ij}(t) \coloneqq \langle \bU^{(\alpha)}_i(t),
J(\bU^{(\alpha)}_j(t))\rangle_{[\alpha]}$, which is still invertible
(since the $\bU^{(\alpha)}_{i_\alpha}$ are linearly independent by
Proposition~\ref{prop:tangent_space}) but no longer symmetric.
The resulting system retains the same structure as~\eqref{eq:DF:MCTDH}
but with all inner products replaced by duality pairings — this is
the general Banach-space form of the MCTDH equations.
\end{remark}

\subsection*{The Hartree manifold}

\begin{example}[Time-dependent Hartree method]
\label{ex:hartree}
For $\fr = \mathbf{1} = (1, \ldots, 1)$, the Tucker manifold
$\Tucker{\mathbf{1}}{\VD}$ consists of elementary tensors
$\bv = \lambda \bigotimes_\alpha v_\alpha$ ($v_\alpha \in V_\alpha$,
$\lambda \in \RR$), called the \emph{Hartree manifold}~\cite{Lubich2008}.
The Dirac--Frenkel ODE~\eqref{eq:DF_tucker} on the Hartree manifold
gives the \emph{time-dependent Hartree equations}. In a Hilbert space
with $\|v_\alpha\|_\alpha = 1$ (unit sphere constraint):
\begin{align*}
    \dot{\lambda}(t) &= \langle \bigotimes_\alpha v_\alpha(t),
        \bF(t, \bv_r(t)) \rangle_D, \\
    \lambda(t)\,\dot{v}_\alpha(t) &= P_{W_\alpha(t)}\bigl[
        (\Id{\alpha} \otimes \langle \cdot, \bigotimes_{\beta \neq \alpha}
        v_\beta(t) \rangle)\bF(t, \bv_r(t))\bigr],
\end{align*}
where $P_{W_\alpha(t)}$ is the orthogonal projection onto the
complement of $\spann\{v_\alpha(t)\}$ in $V_\alpha$. This is
the system solved numerically by MCTDH.
\end{example}

%------------------------------------------------------------
\section{Extension to tree-based formats}
\label{sec:DF:treebased}
%------------------------------------------------------------

The extension of Theorem~\ref{thm:DF_tucker} to $\FTv$ is immediate
given the immersion result of Chapter~\ref{chap:TB_manifold}.

\begin{theorem}[Dirac--Frenkel for tree-based formats,
{\cite[Section~5]{FHN2023}}]
\label{thm:DF_treebased}
Let $\TD$ be a dimension partition tree, $\fr \in \ADTD$ such that
$\FTv$ is a proper set of tree-based tensors. Assume $\VDnorm$ is
reflexive and strictly convex, and that \condINJ\ holds at each
level of $\TD$. Let $\bv_0 \in \FTv$.
\begin{enumerate}
\item[\textup{(Existence)}] There exists a differentiable curve
    $t \mapsto \bv_r(t)$ in $\FTv$ satisfying
    \begin{equation}
    \label{eq:DF_TBT}
        \dot{\bv}_r(t) = \Pmetric{\bv_r(t)}\bigl(\bF(t, \bv_r(t))\bigr),
        \qquad \bv_r(0) = \bv_0,
    \end{equation}
    where $\Pmetric{\bv_r(t)}$ is the metric projection onto
    $T_{\bv_r(t)}\mathfrak{i}(\mathbb{T}_{\bv_r(t)}(\FTv))
    = \bigcap_k \ZD[\PkT{k}{\TD}]{\bv_r(t)}$.
\item[\textup{(Characterisation)}] The curve satisfies~\eqref{eq:DF_TBT}
    if and only if $\dot{\bv}_r(t) \in \bigcap_k \ZD[\PkT{k}{\TD}]{\bv_r(t)}$
    and
    \begin{equation}
    \label{eq:DF_TBT_char}
        \|\dot{\bv}_r(t) - \bF(t, \bv_r(t))\|_D
        = \min_{\dot{\bv} \in \bigcap_k \ZD[\PkT{k}]{\bv_r(t)}}
        \|\dot{\bv} - \bF(t, \bv_r(t))\|_D.
    \end{equation}
\end{enumerate}
\end{theorem}

\begin{proof}
The proof is identical to that of Theorem~\ref{thm:DF_tucker}, with
$\Tucker{\fr}{\VD}$ replaced by $\FTv$ and $\ZD{\bv_r(t)}$ replaced
by $\bigcap_k \ZD[\PkT{k}]{\bv_r(t)}$. The key inputs are:
(i) $\FTv$ is a $\mathcal{C}^\infty$-Banach manifold
(Theorem~\ref{thm:TBT_Banach_Manifold}); (ii) its tangent space is
a split subspace of $\VDnorm$ (Theorem~\ref{thm:TBT_immersion}); and
(iii) the metric projection onto a split subspace in a reflexive
strictly convex space is well-defined and unique.
\end{proof}

\begin{remark}[Algorithmic interpretation]
\label{rem:DF_algorithms}
Theorem~\ref{thm:DF_treebased} provides the mathematical foundation
for a class of algorithms that integrate tensor-format ODEs by evolving
the transfer tensors $\{C^{(\alpha)}\}_{\alpha \in \TD}$ and the leaf
basis vectors $\{u^{(j)}_{i_j}\}_{j \in \Leaves}$ along the
Dirac--Frenkel trajectories. For the TT format, this is the starting
point of the \emph{projector-splitting integrator} of Lubich, Oseledets,
and Vandereycken~\cite{LubichOseledetsVandereycken2015}. For general
tree-based formats, the corresponding algorithm was introduced by
Ceruti, Lubich, and Walach~\cite{CerutiLubichWalach2020}.
\end{remark}

%------------------------------------------------------------
\section{Kinematic equations for tree-based formats}
\label{sec:DF:kinematic_TBT}
%------------------------------------------------------------

The derivation of the kinematic system extends naturally to tree-based
formats via the intersection structure of the tangent space.

\subsection*{Parametrisation of the tangent space}

For $\bv_r(t) \in \FTv$, the tangent space is
$\bigcap_{k=1}^{\dep{\TD}} \ZD[\PkT{k}{\TD}]{\bv_r(t)}$
(Proposition~\ref{prop:TBT_tangent}). A natural parametrisation
is provided by the tree-based chart coordinates:
\begin{itemize}
\item For each leaf $j \in \Leaves$: a basis
    $\{u^{(j)}_{i_j}(t)\}_{i_j=1}^{r_j}$ of $\Umin{j}{\bv_r(t)}$.
\item For each internal vertex $\alpha \in \TD \setminus (\{D\} \cup \Leaves)$:
    a basis $\{U^{(\alpha)}_{i_\alpha}(t)\}_{i_\alpha=1}^{r_\alpha}$
    of $\Umin{\alpha}{\bv_r(t)} \subset \bigotimes_{\beta \in \Sons{\alpha}}
    \Umin{\beta}{\bv_r(t)}$.
\item The transfer tensors $C^{(\alpha)}(t)$ at each internal vertex,
    encoding the coefficients of $\Umin{\alpha}{\bv_r(t)}$ in
    $\bigotimes_{\beta \in \Sons{\alpha}} \Umin{\beta}{\bv_r(t)}$.
\end{itemize}

A tangent vector $\dot{\bv}_r(t) \in \bigcap_k \ZD[\PkT{k}]{\bv_r(t)}$
decomposes at each level $k$ according to
Proposition~\ref{prop:tangent_space}:
\begin{equation}
\label{eq:DF:TBT_tangent_decomp}
    \dot{\bv}_r(t)
    = \sum_{\alpha \in \PkT{k}{\TD}} \sum_{(i_\alpha)}
      \dot{C}^{(\alpha)}_{(i_\alpha)}(t)
      \bigotimes_{\alpha \in \PkT{k}{\TD}} u^{(\alpha)}_{i_\alpha}(t)
    + \sum_{\alpha \in \PkT{k}{\TD}} \sum_{i_\alpha}
      \dot{u}^{(\alpha)}_{i_\alpha}(t) \otimes
      \bU^{(\alpha)}_{i_\alpha}(t),
\end{equation}
where the decomposition is independent of the level $k$ by
Proposition~\ref{prop:laplacian_exp}.

\subsection*{The kinematic system}

The Dirac--Frenkel condition~\eqref{eq:DF:variational_banach} tested
against basis elements of $\bigcap_k \ZD[\PkT{k}]{\bv_r(t)}$ yields
a system indexed by the tree structure.

\medskip
\noindent\textbf{Equations for the root transfer tensor.}
Testing against $\xi = \bigotimes_{\alpha \in \PkT{1}{\TD}}
u^{(\alpha)}_{j_\alpha}(t)$ (for each multi-index
$(j_\alpha)_{\alpha \in \PkT{1}{\TD}}$):
\begin{equation}
\label{eq:DF:TBT_root}
    \dot{C}^{(D)}_{(j_\alpha)}(t)
    = \bigl\langle
        \bigotimes_{\alpha \in \PkT{1}{\TD}} u^{(\alpha)}_{j_\alpha}(t),\;
        J\bigl(\bF(t,\bv_r(t))\bigr)
    \bigr\rangle
    + \text{(gauge correction terms)}.
\end{equation}

\noindent\textbf{Equations for the internal transfer tensors.}
For each internal vertex $\alpha \in \TD \setminus (\{D\} \cup \Leaves)$,
testing against $\xi = w^{(\alpha)} \otimes \bU^{(\alpha)}_{j_\alpha}(t)$
with $w^{(\alpha)} \in \Wmin{\alpha}{\bv_r(t)}$:
\begin{equation}
\label{eq:DF:TBT_internal}
    P^{(\alpha)}_W(t)
    \Bigl[
        \sum_{i_\alpha} F^{(\alpha)}_{i_\alpha j_\alpha}(t)\,
        U^{(\alpha)}_{i_\alpha}(t)
        - M^{(\alpha)}(t)\, \dot{U}^{(\alpha)}_{j_\alpha}(t)
    \Bigr] = 0,
\end{equation}
where $F^{(\alpha)}_{i_\alpha j_\alpha}(t)$ is the mean-field matrix
at the $\alpha$-level (defined analogously to~\eqref{eq:DF:meanfield}),
and $M^{(\alpha)}(t)$ is the mass matrix of the $\bU^{(\alpha)}_{i_\alpha}$.

\noindent\textbf{Equations for the leaf basis vectors.}
For each leaf $j \in \Leaves$ and $\dot{u}^{(j)}_{i_j}(t) \in
\Wmin{j}{\bv_r(t)}$, the equation has the same form
as~\eqref{eq:DF:basis_explicit}.

\begin{remark}[Recursive structure and the projector-splitting integrator]
\label{rem:TBT_kinematic_recursive}
The kinematic system~\eqref{eq:DF:TBT_root}--\eqref{eq:DF:TBT_internal}
has a recursive structure matching the tree $\TD$: the equation at each
internal vertex $\alpha$ involves only the transfer tensors of $\alpha$
and its children $\Sons{\alpha}$. This recursive structure is precisely
what the \emph{projector-splitting integrator}
of~\cite{CerutiLubichWalach2020} exploits: it solves the kinematic
system level by level, from the root down to the leaves (forward sweep)
and back (backward sweep), at cost $O(\dep{\TD} \cdot dr^3)$ per
time step.

For the TT format ($\TD = \TDtt$, $\dep{\TD} = d-1$), the system
reduces to $d$ coupled equations — one per bond — each involving only
adjacent components. This is the TT version of the projector-splitting
integrator of~\cite{LubichOseledetsVandereycken2015}:
\begin{equation}
\label{eq:DF:TT_splitting}
    \dot{C}^{(k)}(t)
    = P^{(k)}_W(t)\bigl[M^{(k)}(t)^{-1}
      F^{(k)}(t)\bigr],
    \quad k = 1, \ldots, d,
\end{equation}
where $C^{(k)}(t)$ is the $k$-th TT component and $F^{(k)}(t)$
is the mean-field force from the left and right environments.
\end{remark}

%------------------------------------------------------------
\section{Historical notes}
\label{sec:DF:history}
%------------------------------------------------------------

The Dirac--Frenkel variational principle was proposed independently by
Dirac~\cite{Dirac1930} and Frenkel~\cite{Frenkel1934} in 1930 and 1934
for the time-dependent Schrödinger equation in Hilbert space. The
formulation used here — projection of the vector field onto the tangent
space of a manifold — is that of McLachlan et al.\ and Lubich.

Koch and Lubich~\cite{KochLubich2007} proved existence and uniqueness
of solutions to the Dirac--Frenkel ODE on Tucker manifolds in
finite-dimensional Hilbert spaces, and used it to analyse the MCTDH
method. The extension to infinite-dimensional Banach spaces
(Theorem~\ref{thm:DF_tucker}) was established in~\cite{FHN2019},
where the key difficulty — showing that the tangent space $\ZD{\bv_r(t)}$
is a split subspace of $\VDnorm$, without the Hilbert space structure —
was resolved by the immersion theorem (Theorem~\ref{thm:tucker_immersion}).

The extension to tree-based formats (Theorem~\ref{thm:DF_treebased})
is the main application result of~\cite{FHN2023}: having established
the geometry of $\FTv$ in Chapter~\ref{chap:TB_manifold}, the
Dirac--Frenkel principle follows with no additional effort.

